### 2022-05-05 Rmarkdown/paper-si.Rmd
upload paper

---
title: "Supplementary information for *Comparing corrosion control treatments using a robust Bayesian generalized additive model*"
author: "Benjamin F. Trueman^§^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, NS, CAN B3H 4R2    \n&nbsp;    \n&nbsp;^†^EPCOR Water Services, Edmonton, AB, CAN"
output:
  word_document:
    keep_md: no
    reference_docx: StylesTemplate.docx
  html_document: default
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper-si.Rmd")
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("stringr")

ortho_time <- readr::read_csv(here::here("data-clean/ortho-timeline.csv"))
flows <- readr::read_csv(here::here("data-clean/flow-schedule.csv"))
chlorine <- readr::read_csv(here::here("data-clean/chlorine-timeline.csv"))
wq <- readr::read_csv(here::here("data-clean/wq-summary.csv"))
```

This document contains 5 figures, 4 tables, and 10 pages.

```{r table-s1}

options(knitr.kable.NA = ".")

ortho_time %>% 
  knitr::kable(caption = "**Table S1.** Orthophosphate dosing schedule.")

```

```{r table-s2}
chlorine %>% 
  knitr::kable(caption = "**Table S2.** Total chlorine in pipe section influent (mg L^-1^).")
```

```{r table-s3}

flows %>% 
  rename_all(str_to_sentence) %>% 
  knitr::kable(caption = "**Table S3.** Pipe array flow schedule; sampling occurred at T4, and FY-001, -002, and -003 are defined in Figure S1.")

```

```{r table-s4}

wq %>% 
  relocate(Units, .after = Analyte) %>% 
  mutate(Location = if_else(Location == "Treated water", "Treated", Location)) %>% 
  knitr::kable(caption = "**Table S4.** Water quality summary.")

```

```{r figure-s1, fig.cap="**Figure S1.** Schematic of the pipe loop system."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s1.png"))
```

```{r figure-s2, fig.cap="**Figure S2.** Choice of prior did not appear to have a strong impact on the posterior distribution."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s2.png"))
```

Figures S3--4 and S6--7 summarize the parameters of models fitted to observed and simulated data, respectively, including all of the spline coefficients (the weights applied to the basis functions). The statistic $\widehat{R}$ is a convergence metric; at convergence $\widehat{R}$ is equal to 1, and an $\widehat{R}$ less than 1.1 is usually taken to represent approximate convergence [@gelman_bayesian_2014]. Here, all $\widehat{R}$ values were less than 1.05. The effective sample size $N_{eff}$ measures the amount by which autocorrelation in the MCMC chains increases uncertainty in the parameter estimates. That is, $N_{eff}$ has the same estimation power as the $N$ autocorrelated samples. An effective sample size greater than 100 per chain is considered adequate. Here, effective sample sizes for all parameters exceeded this threshold.

```{r figure-s3, fig.cap="**Figure S3.** Summary of model parameters and convergence statistics (dissolved lead model). Medians calculated from the posterior distributions of the **(a)** spline coefficients, **(b)** other parameters (as labelled), **(c)** random intercepts, and **(d)** standard deviations of the group-level multiyear splines. Error bars span the middle 95% of each distribution. $\\widehat{R}$, a convergence statistic, is mapped to point colour and size."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s3.png"))
```

```{r figure-s4, fig.cap="**Figure S4.** Summary of model parameters and convergence statistics (particulate lead model). Medians calculated from the posterior distributions of the **(a)** spline coefficients, **(b)** other parameters (as labelled), **(c)** random intercepts, and **(d)** standard deviations of the group-level multiyear splines. Error bars span the middle 95% of each distribution. $\\widehat{R}$, a convergence statistic, is mapped to point colour and size."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s4.png"))
```

```{r figure-s5, fig.cap="**Figure S5.** In the model fitted to simulated data, **(a)** Seasonal component of the trend in dissolved lead. **(b)** Global multiyear component of the trend in dissolved and particulate lead. **(c)** Group-level (i.e., pipe-specific) multiyear trends in dissolved and particulate lead, representing the deviations of each time series from the global multiyear trend. In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue. All plots are shown on the transformed scale and centered."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s5.png"))
```

```{r figure-s6, fig.cap="**Figure S6.** Summary of model parameters and convergence statistics (dissolved lead model fitted to simulated data). Medians calculated from the posterior distributions of the **(a)** spline coefficients, **(b)** other parameters (as labelled), **(c)** random intercepts, and **(d)** standard deviations of the group-level multiyear splines. Error bars span the middle 95% of each distribution. $\\widehat{R}$, a convergence statistic, is mapped to point colour and size."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s6.png"))
```

```{r figure-s7, fig.cap="**Figure S7.** Summary of model parameters and convergence statistics (particulate lead model fitted to simulated data). Medians calculated from the posterior distributions of the **(a)** spline coefficients, **(b)** other parameters (as labelled), **(c)** random intercepts, and **(d)** standard deviations of the group-level multiyear splines. Error bars span the middle 95% of each distribution. $\\widehat{R}$, a convergence statistic, is mapped to point colour and size."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s7.png"))
```

```{r figure-s8, fig.cap="**Figure S8.** Simulated residuals, grouped by orthophosphate dosing protocol and pipe material. Points represent medians summarizing 4000 draws from the posterior, and error bars span the 95% credible intervals."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-s8.png"))
```

# References

### 2022-05-05 Rmarkdown/paper.Rmd
upload paper

---
title: "Comparing corrosion control treatments using a robust Bayesian generalized additive model"
author: "Benjamin F. Trueman^§^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"
date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, NS, CAN B3H 4R2    \n&nbsp;    \n&nbsp;^†^EPCOR Water Services, Edmonton, AB, CAN"
output:
  word_document:
    keep_md: no
    reference_docx: StylesTemplate.docx
  html_document: default
bibliography: references.bib
csl: environmental-science-and-technology.csl
---

```{r setup, include=FALSE}
here::i_am("Rmarkdown/paper.Rmd")
knitr::opts_chunk$set(echo = FALSE)
library("dplyr")
library("stringr")
model_in <- readr::read_csv(here::here("data-clean/model_in.csv"))
fig7_summ <- readr::read_csv(here::here("data-clean/figure-7-summary.csv"))
```

# Abstract

Pipe loop studies are used routinely to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain an important tool for water quality management. But they are inherently complex and the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common features, the latter due to repeatedly measuring the effluent from test pipes. Popular statistical tests, such as the Student *t* or rank-sum tests, are often inadequate descriptions of the data. Here, we propose an approach to statistical analysis of pipe loop data that accommodates many of these difficult-to-model characteristics: a robust Bayesian generalized additive model with continuous-time autoregressive errors. Our model facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers. It is well-suited to describing nonlinear trends without overfitting, and it accounts for reduced information content due to autocorrelation. We demonstrate the model using an example pipe loop study comprising four years of data, multiple pipe configurations, and multiple orthophosphate dosing protocols. We compare the experimental treatments, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term. We also find that---consistent with previous work---galvanic corrosion yields relatively high particulate lead release, especially at higher orthophosphate doses. An advantage of the Bayesian approach we adopt here is that it yields the full posterior distribution of all parameters and all quantities derived from those parameters, including the model predictions. This means that analysts can construct any comparison that is relevant to the study goals, without the need to rely on a particular statistical test.

```{r toc-art}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-toc.png"))
```

# Introduction

Pipe loop studies are used to model drinking water distribution systems, mainly for evaluating corrosion control treatment. The US EPA's Revised Lead and Copper Rule is expected to result in many water utilities conducting pipe loop studies [@estessmargiassi_understanding_2020], and the data they generate will need to be analyzed. But pipe loop study data are inherently complex and there is a real risk of misinterpreting them. The following are just some of the challenges that arise (Figure 1):

(1) Non-linear time trends and seasonal variation. These features can be difficult to describe without over- or under-fitting. They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995]. 
(2) Left-censored values (measurements below the detection limit), which many popular statistical tests are not designed to handle.
(3) Autocorrelation, or memory effects. Autocorrelation means the current state of the system depends on its history. For example, the effect of a perturbation to corrosion scale is evident for some time after the fact. From a statistical perspective, autocorrelation makes parameter estimates more uncertain [@helsel1992statistical] by reducing the effective sample size (i.e., the information content of the sample) [@gelman_bayesian_2014]. It is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time.
(4) An irregular sampling frequency, which makes it more difficult to estimate the degree of autocorrelation.
(5) Extreme lead or copper concentrations due to particle release. Extreme values are often inconsistent with the common assumption of  Gaussian residuals.

```{r figure-1, fig.cap="**Figure 1.** An example time series from the study described in this paper; labels 1--5 represent features of the series that represent statistical modeling challenges."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-1.png"))
```

All of the above present barriers to reliable statistical inference, but left-censoring and autocorrelation can be particularly challenging. Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values. Neither approach is advisable because both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012]. Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014]. As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel1992statistical]. The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel1992statistical]. That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014]. Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data. And while rank-based procedures such as the rank-sum test can accomodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].

Here, we propose a general approach for building statistical models of pipe loop study data that addresses all of these challenges. The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of four components:

1. A global---possibly nonlinear---trend describing variation in the parameter of interest over time across all pipes.
2. A set of group-level trends that describe the deviations from the global trend particular to the individual pipes.
3. A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature. (This component may not be necessary if there is no distinct seasonality.)
4. An autocorrelated error term.

We demonstrate this approach using an example pipe loop dataset representing four years of data, multiple pipe configurations, and multiple orthophosphate dosing protocols. Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary. An advantage of our Bayesian approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values. This means that we can construct any comparison we wish to investigate, without the need to decide on a particular statistical test. Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in lead concentrations over time.

# Methods

## Model distribution system

The model system comprised three pipe arrays (Figure S1), each with five pipes:

1. A 9.4 m, 1/2" ID lead pipe (Pb #1).
2. A duplicate of (1), Pb #2.
3. A 3 m, 1/2" ID lead pipe coupled galvanically to a 6.4 m, 1/2" ID copper pipe.
4. A 9.4 m, 1/2" ID copper pipe with twenty-two 50/50 lead-tin solder joints ("Cu/Pb-Sn") (not discussed here).
5. A 9.4 m, 1/2" nominal internal diameter (ID) copper pipe (not discussed here).

All plumbing was new at the time of installation, and lead pipes were acquired from Canada Metals Eastern (ASTM B29, 99.94% pure lead). Each array was preceded in the flow-path by a 3.6 m, 3" ID ductile iron pipe, and system pressure was approximately 40 psi.

The model system was conditioned from Sept. 2017 to Mar. 2018 using treated water from the E.L. Smith Water Treatment Plant in Edmonton, AB, CAN. Afterward, each array received a different dose of phosphoric acid: 0 mg P L^-1^ (increasing to 0.5 mg L^-1^ in Jan. 2019), 1.0 mg L^-1^, and 2.0 mg L^-1^ (decreasing to 0.75 mg L^-1^ in Apr. 2019). The orthophosphate dosing schedule is summarized in Table S1. Beginning in Nov. 2020, influent to the model system was partially dechlorinated by blending granular activated carbon filtrate with chlorinated water upstream of the pipe sections; the resulting total chlorine concentrations are summarized in Table S2. In November 2020, 30 cm sections of lead pipe were cut from the galvanic lead-copper pipes for supplementary analysis. These sections were replaced with braided PVC hoses.

Flow through the pipe sections was programmed to approximate typical use and is summarized in Table S3. Samples were collected weekly until Mar. 2021, after which they were collected every two weeks. At each sampling event, 1 L was collected from each pipe after a 6-hour stagnation period. Samples were collected at a flow rate of approximately 3 L min^-1^.

## Water quality analysis

Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively. An aliquot of each sample was filtered at 0.45 µm using a teflon filter and a vacuum-filter apparatus. Temperature was measured using an infrared thermometer, total chlorine [@sm_4500_cl_g] and orthophosphate [@sm_4500_p_e] were measured using a spectrophotometer, and pH was measured using a benchtop meter. Water quality---including alkalinity [@sm_2320b], true colour [@sm_2120c], conductivity [@sm_2510b], fluoride [@sm_4500], total hardness [@sm_2340], and turbidity [@sm_2130b] ---is summarized in Table S4. Colour was quantified after filtration at 0.45 µm, using mixed cellulose ester membrane filters.

```{r figure-2, fig.cap="**Figure 2. (a)** Constant, linear, and generalized additive models (GAMs) fitted to an example (log-transformed) time series from the dataset, along with the residuals from each model. **(b)** Pearson correlation (*r*) between the residuals $\\epsilon$ at time *t* and *t-1*. **(c)** Pearson correlation between the residuals $\\epsilon$ at time *t* and *t-k*, where *k* is the lag and ranges from 0--21. The grey shaded regions represent $\\pm1.96/\\sqrt{n}$, or an approximate 95% confidence bound on the autocorrelation function of an uncorrelated series."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-2.png"))
```

## Model selection

Comparisons of corrosion control treatments are often aided by a statistical model. Typical choices in pipe loop studies include constant (*e.g.*, *t*-test) and linear models. But when applied to an example time series from our dataset (Figure 2a), both of these models violate a key assumption: that residuals are independent.

Lack of independence is apparent in Figure 2b: the residual of either model at time *t*, $\epsilon_t$, is correlated with the residual at time *t-1*, $\epsilon_{t-1}$. This concept is extended to include the residual at time *t-k*, $\epsilon_{t-k}$, where *k = 1...21* to yield the autocorrelation function (Figure 2c). Independent residuals yield autocorrelation functions that hover around zero, whereas correlated residuals typically yield autocorrelation functions that either decay or oscillate. We assume here that the example time series is a regular sequence, but in this case, that is only an approximation, and we relax it in the next section.

```{r figure-3, fig.cap="**Figure 3.** The unweighted (cubic) basis functions of *x* on which *log*([Pb]) is regressed, and the same functions after weighting by the model coefficients. The (centered) model fit, or the sum of the weighted basis functions (equation 2), is superimposed in black."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-3.png"))
```

The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2). This is because it accounts for non-linear patterns that the simpler models ignore. GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).

$$(1)~~y = \beta_{0} + f(x_t) + \epsilon_t$$ 
where $f(x_t)$ is a smooth function of $x_t$,

$$(2)~~f(x_t) = \sum_{j=1}^kb_j(x_t)\beta_j$$
$b_j(x_t)$ are the functions of $x_t$ on which $y$ is regressed, $\beta_j$ are the model coefficients, and $\epsilon_t$ is the error term. We fitted the example GAM in Figures 2--3 using the R package *mgcv*, which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.

### Residual autocorrelation

Often, even a flexible model---such as a GAM---fails to account for residual autocorrelation in a time series, in which case it is possible to relax the independence assumption and model the residuals as an *autoregressive* process. A very common model is the first-order autoregressive (AR(1)) process:

$$(3)~~\epsilon_t = \phi\epsilon_{t-1} + \omega_t$$
where $\epsilon_t$ is the residual at time $t$, $\phi$ is the autoregressive parameter, and $\omega_t$ is an independent draw from a probability distribution.

First-order autocorrelation can be approximated as the correlation between the lagged and unlagged residuals, $\epsilon_t$ and $\epsilon_{t-1}$. A reasonable estimate of $\phi$ for the linear model residuals in Figure 2b, then, would be *r*. Unmodeled positive autocorrelation, as in Figure 2b-c, overstates the precision of parameter estimates, and so it should be either eliminated via model selection or taken into account explicitly. We chose to analyze the full length of each time series---including the conditioning period---to best estimate the autocorrelation structure.

While equation (3) is limited to time series that are regularly observed, it can be generalized to an irregular sequence of observations as a continuous-time first-order autoregression [@simpson_modelling_2018]. This more general model has the following autocorrelation function:

$$(4)~~h(s, \phi) = \phi^s, s\geq0, \phi\geq0, \phi\leq1 $$

In equation $(4)$, $s$ represents the separation in time between successive observations. Autocorrelation, then, decays exponentially in time; if two observations are spaced far enough apart, they are approximately independent. Given the irregular spacing of observations in our data, we modeled autocorrelation using equation $(4)$.

## Model description

To compare corrosion control treatments using data from the model distribution system, we fitted separate GAMs with continuous-time first-order autocorrelated errors to the particulate (>0.45 µm) and dissolved (<0.45 µm) lead time series. Particulate lead was determined as the difference between total and dissolved lead. Particulate lead concentrations less than the MDL of 0.2 µg L^-1^ were left-censored, and particulate lead concentrations representing the difference between an observed (total) and left-censored (dissolved) concentration were interval-censored. The lower bound of the interval was set as the difference between the total lead concentration and the detection limit and the upper bound as the total lead concentration. Given their impractically high censoring rates (up to 96%), time series representing effluent from soldered copper pipe sections were omitted.

Model code was written in Stan [@stan2021; @rstan], a C++ package for Bayesian inference using Markov chain Monte Carlo (MCMC) methods. We generated a code template using the R package `brms` [@brms; @brms_adv] and modified it to specify the continuous-time autocorrelation structure defined in equation $(4)$. We used a suite of common R packages [@tidyverse; @nada; @assertr; @janitor; @datatable], and bespoke functions used to fit the model and process the results are documented on GitHub [@bgamcar1]. The data and R code necessary to reproduce the analysis are available in a separate repository [@compare_cct].

```{r figure-4, fig.cap="**Figure 4. (a)** Seasonal component of the trend in dissolved lead. **(b)** Global multiyear component of the trend in dissolved and particulate lead. **(c)** Group-level (i.e., pipe-specific) multiyear trends in dissolved and particulate lead, representing the deviations of each time series from the global multiyear trend. In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue. All plots are shown on the transformed scale and centered."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-4.png"))
```

The models included one-dimensional global smooths describing multiyear variation in lead across all time series (Figure 4b). The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a). Both models included group-level smooths describing multiyear variation in lead release from each pipe (Figure 4c), a heirarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019]. The multiyear smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv]. To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.

Time series representing lead release to drinking water tend to feature extreme values caused by particulate or colloidal lead. These can severely bias parameter estimates unless they are accounted for. And since our dataset was impacted by extreme lead release events, we used a Student *t* likelihood in place of the usual Gaussian likelihood to limit the influence of outliers on the posterior distribution [@brms; @brms_adv].

The transformed data, $y_t$, were modeled as follows,

$$
(5)
\begin{align}
y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu) \\ 
y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu) \\
y_t | censored_t = 2 \sim G(\mu_t,\sigma,\nu) \\
\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s(y_{t-s} - \mu_{t-s}) \\
\sigma \sim T(0, 2.5, 3) \\
\nu \sim Gamma(2,0.5) \\
\phi \sim N(0.5, 0.25) \\
\end{align}
$$

where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$. The autocorrelation coeffient $\phi^s$ is defined in equation $(4)$.

Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that a random variable $X$ is at most $x$) [@stan2021; @mcelreath_statistical_2016]. Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$. The local intercepts $\alpha_{pipe_i}$ are defined as follows:

$$
(6)
\begin{align}
\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha) \\
\alpha \sim T(0.1, 2.5, 3) \\
\sigma_{\alpha} \sim T(0, 2.5, 3) \\ 
\end{align}
$$

And the $f_j(t)$ are smooth spline functions that comprise the GAM, each with the following form:

$$
(7)
\begin{align}
f_j(t) = X_j\beta_j + Z_jb_j \\
\beta_j \sim T(0, 2.5, 3) \\
b_j \sim N(0, \sigma_b) \\
\sigma_b \sim N(0, 0.5) (seasonal~smooth) \\
\sigma_b \sim T(0, 2.5, 3) (all~other~smooths) \\ 
\end{align}
$$

where $Z_j$ and $X_j$ are design matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function of the global multiyear trend). The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooths [@mgcv]. And while the chosen priors did help with model convergence, they did not exert a strong influence on the posterior (Figure S2); the likelihood---representing `r nrow(model_in)` observations in each model---was overwhelmingly influential. A summary of the fitted models, including convergence statistics and effective sample sizes, is provided in Figure S3 and S4.

### Estimating derivatives

We estimated the local slope of the global multiyear smooths using a finite difference method. Specifically, we generated 4000 posterior predictions of the smooth trend along a regular sequence spanning the date range of the data. This was repeated after adding $\delta = 0.001$ (~$1/3$ of a day) to each value in the sequence. The difference between the grid of posterior predictions evaluated at $t$ and $t+\delta$, divided by $\delta$, approximates the instantaneous rate of change after accounting for seasonal variation, inter-pipe variation, and autocorrelation.

### Model parameters and diagnostics

Residuals from the two models were approximated well by the estimated residual error distribution (Figure 5a). Furthermore, models fitted to a draw from the posterior predictive distribution---that is, to simulated observations---yielded similar parameters to those of the original models (Figures S5--7).

The continuous-time autocorrelated error term was successful in capturing residual first-order autocorrelation, estimated in Figure 5b using the Spearman rank correlation coefficient as a measure that is robust against outliers. Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset. Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation. The distribution of the autocorrelation coefficient $\phi$ is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.

```{r figure-5, fig.cap="**Figure 5. (a) ** Histograms representing the median of 4000 residual draws from the models; the theoretical densities are calculated from the posterior medians of the distributional parameters (i.e., $\\sigma$, $\\nu$) from each model. **(b)** Density of first-order autocorrelation, estimated using the Spearman rank correlation coefficient and 4000 residual draws from the model including and excluding the continuous-time first-order autocorrelation term. Estimates were generated using an augmented dataset where censored observations were imputed via posterior prediction. **(c)** Posterior distributions of $\\phi$, the autocorrelation coefficient. **(d)** Pareto *k* shape parameters, which estimate the influence of each observation on the posterior distribution."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-5.png"))
```

The posterior predictive distribution was also robust against extreme values (Figure 5d): the Pareto-*k* values shown here represent the influence of individual observations on the posterior distribution, and values below 0.7 are considered moderately influential, yielding reliable estimates of out-of-sample predictive accuracy (see *Comparing corrosion control treatments* below) [@loo; @loo_paper]. Here, all Pareto-*k* values were below 0.5.

# Results

## Comparing corrosion control treatments

Time series of particulate and dissolved lead featured prominent negative trends in the first year of operation followed by an extended period of relative stability. The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a). To estimate the time-varying mean representing each series, we modeled lead as the sum of a global multiyear smooth, a global seasonal smooth (dissolved lead only), a group-level multiyear smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).

Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$. The effect of the autocorrelation term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe copuples in Nov. 2020 ("Pb-Cu", Figure 6). While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4). That is, extreme and sustained deviations from typical behavior are generally captured by the error term.

Accounting for autocorrelation also improved the predictive accuracy of the models. Approximate leave-one-out cross-validation suggests that models with continuous-time autoregressive errors yielded better predictions than otherwise identically-specified models with independent errors. Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016]. The difference between the expected log pointwise predictive densities of the dissolved and particulate lead models, respectively, was 206 and 47. In both comparisons, the autoregressive error model was more accurate.

```{r figure-6, fig.cap="**Figure 6.** Time series of dissolved and particulate lead representing three orthophosphate dosing protocols (columns) and two pipe configurations (rows; the full lead pipe configuration---Pb #1--2---is duplicated). The expected value of the posterior predictive distribution (*i.e.*, $\\mu_t$ in Equation 5) is summarized by the median (blue and red lines) and the 2.5--97.5^th^ percentiles (red/blue shaded regions). The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections, the yellow shaded bands represent brief periods when orthophosphate dosing was interrupted, and black ticks at the top of each plot represent values outside the plot limits. Other significant disturbances are labeled, as described in the section titled *Model distribution system*"}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-6.png"))
```

Posterior predictions from the models---$\mu_t$ in Equation $(5)$---can be compared to contrast the orthophosphate dosing protocols. Here we compared the 0--0.5 mg P L^-1^ with the 1 mg P L^-1^ and the 1 mg P L^-1^ with the 2--0.75 mg P L^-1^ dosing protocols over time by considering their ratios.

The three protocols yielded distinct lead release distributions over most of the study, and the difference between the lowest orthophosphate dose (0.5 mg P L^-1^) and the other two is particularly apparent (Figure 7). The ratio of dissolved lead concentrations in the absence of orthophosphate to those representing the 1 mg P L^-1^ dose peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a); the ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^, but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 for the majority of the study. This pattern is consistent with expectations based on geochemical solubility models [@schock1996corr].

The ratio comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b). As a comparison of either particulate or lead concentrations, it was less than one over much of the first part of the series and greater than one over much of the latter part. The transition occurred a short time after orthophosphate was decreased from 2--0.75 mg P L^-1^, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose, likely as particles and small colloids. Meanwhile, this dose presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.

```{r figure-7, fig.cap="**Figure 7.** Ratios of posterior predictions comparing **(a)** 0--0.5 with 1 mg P L^-1^ (numerator and denominator, respectively) and **(b)** 1 with 2--0.75 mg P L^-1^. The bold black/red line represents the median of the posterior distribution and the shaded grey region spans the 95% credible interval; black and red indicate portions of the series where the interval included and excluded, respectively, a ratio of 1 (i.e., equality). A random sample of 200 posterior draws is superimposed in blue."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-7.png"))
```

## Identifying periods of significant change

Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the duration of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant. This general pattern is represented in our model by the global multiyear trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature). The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term. Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a). Both fractions exhibited more modest changes in subsequent years that are difficult to attribute, but they may be a response to changes in water quality over the course of the study.

```{r figure-8, fig.cap="**Figure 8. (a) ** The global multiyear smooth and its estimated first derivative. A random sample of 200 smooths/derivatives is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions. The median of the global trend is shaded red when the 95% credible interval around the first derivative does not include zero. **(b--c)** Iron and water temperature, both measured in pipe rack influent, exhibited comparable seasonal patterns to lead, the latter described by the seasonal term in the GAM. Ticks at the top and bottom edges of the plot in **(b)** represent anomalously high and left-censored iron concentrations, respectively."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-8.png"))
```

The variation in dissolved lead captured in Figure 8a neglects seasonality. And while the seasonal component of the model (Figure 8b) is described as a function of ordinal day, this is a proxy for the physical and chemical inputs that are expected to have caused it. The true seasonal relationship is probably complex and difficult to determine from the study data, but both temperature and influent iron concentration varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016].

## Estimating the particulate fraction

Over the duration of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9). Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible. The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018].

```{r figure-9, fig.cap="**Figure 9.** The particulate fraction of total lead over time, grouped by pipe and orthophosphate dosing protocol, and estimated from the posterior distributions of the models fitted to particulate and dissolved lead time series."}
knitr::include_graphics(here::here("Rmarkdown/figures/figure-9.png"))
```

# Conclusion

Pipe loop study data can be difficult to analyze, owing to non-linear time trends, seasonality, autocorrelation due to repeated measurements of pipe effluent, irregular sampling frequency, extreme particulate metals concentrations, and left-censoring, or values below the detection limit. To address these difficulties, we propose a robust Bayesian generalized additive model with continuous-time first-order autoregressive errors. With this approach, there is no need to impute non-detects, assume linearity, or treat outliers specially. Moreover, the model accounts for the reduced information content in autocorrelated observations. We demonstrate it on an example pipe loop dataset and use the model to compare orthophosphate dosing protocols and analyze time trends. The additive smooth terms comprising the model yield an informative decomposition of pipe loop time series and a more nuanced comparison of corrosion control treatments.

The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the ratio difference between the 0 and 1 mg P L^-1^ dose. The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) yielded the lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^. Consistent with previous work, orthophosphate tended to increase the particulate fraction of total lead. The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring. This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.

Given that pipe loop studies are expected to remain important for drinking water quality management, time series analysis will be  necessary for reliable statistical analysis of the data they generate. Methods that do not account for the effect of repeated measurements---and the reduced information content that results from autocorrelation---are likely to be less useful, and may even be misleading. Methods should be flexible enough to account for nonlinear time trends, and they should be robust against extreme values. And since the goal of pipe loop studies is often to minimize the release of lead and other metals into drinking water, concentrations may fall below the analytical range for a given analyte. The most useful statistical methods, then, will be well-suited to modeling censored data.

# Acknowledgements

This work was supported by Mitacs, through the Mitacs Accelerate Program (Reference # IT23352), and NSERC, through an NSERC/Halifax Water Industrial Research Chair program (Grant # IRCPJ: 349838-16). We thank David Redden and Paul Bjorndahl for providing comments on a draft of the manuscript. We also acknowledge EPCOR's analytical operations team for water quality analysis and their
process development team for operating the pipe loops.

# Associated content

Supplementary information for this article includes additional description of the pipe loop study, a sensitivity analysis of the priors, models summaries, and convergence statistics.

# References

### 2022-05-05 R/check-prior.R

@@ -1,0 +1,49 @@
+
+# prior predictive checks (model is loaded from CSVs by default):
+
+#------------------ setup ------------------
+
+library("bgamcar1")
+source("R/formulas.R")
+library("rstan")
+library("ggplot2")
+library("tibble")
+library("tidyr")
+library("dplyr")
+
+options(mc.cores = parallel::detectCores())
+
+#------------------ inputs ------------------
+
+model_in <- readr::read_csv("data-clean/model_in.csv")
+
+#------------------ models ------------------
+
+knots_yday <- c(0, 1)
+
+# prior simulation:
+
+model_prior_part <- fit_stan_model(
+  "data-clean/prior_part",
+  seed = stan_seed,
+  form_part, model_in, prior_part,
+  save_warmup = FALSE,
+  sample_prior = "only"
+)
+
+model_prior_diss <- fit_stan_model(
+  "data-clean/prior_diss",
+  seed = stan_seed,
+  form_diss, model_in, prior_diss,
+  save_warmup = FALSE,
+  sample_prior = "only",
+  knots = list(date_yday = knots_yday)
+)
+
+# smooths:
+
+conditional_smooths(model_prior_part)
+conditional_smooths(model_prior_diss)
+
+model_prior_part$prior
+model_prior_diss$prior

### 2022-05-05 R/cleaning/model-inputs.R

@@ -1,0 +1,160 @@
+
+# generate model inputs:
+
+library("dplyr")
+library("tidyr")
+library("stringr")
+library("testthat")
+library("assertr", include.only = "verify")
+
+#------------------ functions ------------------
+
+# used for scaling the upper bound of censored intervals:
+
+scale_fun <- function(x, y = NULL, log_trans = FALSE) {
+  if(log_trans) {
+    x <- log(x)
+  }
+  if(is.null(y)) {
+    y <- x
+  }
+  m <- mean(y)
+  sdev <- sd(y)
+  
+  (x - m) / sdev
+  
+}
+
+test_that("scale_fun() returns expected result", {
+  x <- runif(10, 1, 2)
+  t1 <- scale_fun(x, log_trans = TRUE)
+  t2 <- scale(log(x))[,1]
+  t3 <- scale_fun(x, y = log(x), log_trans = TRUE)
+  t4 <- scale_fun(x)
+  t5 <- scale(x)[,1]
+  expect_equal(t1, t2)
+  expect_equal(t2, t3)
+  expect_equal(t4, t5)
+})
+
+#------------------ data cleaning ------------------
+
+pdat <- readr::read_csv("data-clean/pdat.csv")
+
+pdat_clean <- pdat %>%
+  mutate(param = str_replace(param, "Lead, dissolved", "Lead Dissolved")) %>%
+  filter(
+    str_detect(location, "^LP[1-3][1235]$"),
+    param %in% c("Lead", "Lead Dissolved"),
+    !is.na(value)
+  ) %>%
+  transmute(
+    type = str_to_lower(param) %>%
+      str_replace("\\s", "_"),
+    ortho_dose,
+    pipe_material,
+    date,
+    location,
+    value,
+    bdl
+  ) %>%
+  distinct() %>%
+  assertr::verify(value >= 2e-4 | (value == 2e-4 & bdl)) %>%
+  assertr::verify(date > "2017-01-01") %>%
+  assertr::verify(date <= lubridate::today())
+
+pdat_part <- pdat_clean %>%
+  pivot_wider(names_from = type, values_from = c(value, bdl), values_fn = median) %>% 
+  # the median of a censored and uncensored value should be censored at the median:
+  mutate(
+    # code as censored any median-aggregated value where the median is censored 
+    # or the midpoint of a censored and uncensored value:
[…]

### 2022-05-05 R/figures.R

@@ -1,0 +1,1464 @@
+
+# generate figures for "paper.Rmd" and "paper-si.Rmd"
+
+#------------------ setup ------------------
+
+source("R/models.R", echo = TRUE)
+library("forcats")
+library("ggplot2")
+library("dplyr") 
+library("tidyr") 
+library("purrr")
+library("patchwork")
+library("readr")
+library("posterior")
+library("withr", include.only = "with_seed")
+library("imputeTS", include.only = "na_interpolation")
+library("ggdist", include.only = c("stat_halfeye", "median_qi"))
+library("glue", include.only = "glue")
+library("assertr", include.only = "verify")
+library("devtools", include.only = "session_info")
+library("janitor", include.only = "clean_names")
+library("wesanderson", include.only = "wes_palette")
+library("ggh4x", include.only = "facetted_pos_scales")
+library("mgcv", include.only = "gam")
+library("broom", include.only = "tidy")
+library("lubridate", include.only = c("year", "yday"))
+library("ggtext", include.only = c("element_markdown", "geom_richtext"))
+library("tibble", include.only = c("tribble", "rowid_to_column"))
+
+theme_set(theme_bw() + theme(legend.position = "bottom"))
+palette <- wesanderson::wes_palette("Zissou1")
+
+#------------------ load ------------------
+
+pdat <- readr::read_csv("data-clean/pdat.csv")
+pdat_clean <- readr::read_csv("data-clean/pdat_clean.csv")
+pk_vals <- readr::read_csv("data-clean/pareto-k-values.csv")
+
+locations <- model_in %>%
+  distinct(location, ortho_dose, pipe_material)
+
+#------------------ figure 1 ------------------
+
+annotations <- tibble::tribble(
+  ~date, ~value, ~label,
+  "2019-01-01", 0.02, "1. Non-linearity",
+  "2020-10-01", 1e-4, "2. Left-censoring",
+  "2020-03-01", 0.1, "3. Autocorrelation\n(clustering of similar\nvalues in time)",
+  "2018-04-01", .7, "4. Irregular sampling\nfrequency",
+  "2021-06-01", .5, "5. Extreme\nvalues",
+) %>% 
+  mutate(date = as.Date(date))
[…]

### 2022-05-05 R/formulas.R

@@ -1,0 +1,43 @@
+
+# write formulas for use in "R/models.R":
+
+library("brms")
+library("stringr")
+
+stan_seed <- 236 # random seed
+
+# formulas:
+
+ar <- 'ar(time = date_numeric, gr = location)'
+seasonal <- 's(date_yday, bs = "cc")'
+
+form_part_noar <- 'scaled_lead_part | cens(cens_lead_part, y2 = scaled_lead) ~
+    1 + (1 | location) +
+    s(date_numeric) +
+    s(date_numeric, by = location, m = 1)'
+
+form_diss_noar <- paste(
+  str_replace_all(form_part_noar, "lead_part", "lead_dissolved") %>% 
+    str_remove_all(", y2 = scaled_lead"), 
+  seasonal, sep = " + "
+)
+
+form_part <- bf(paste(form_part_noar, ar, sep = " + "))
+form_diss <- bf(paste(form_diss_noar, ar, sep = " + "))
+form_part_sim <- bf(
+  paste(str_remove_all(form_part_noar, ", y2 = scaled_lead"), ar, sep = " + ")
+)
+
+form_part_noar <- bf(form_part_noar)
+form_diss_noar <- bf(form_diss_noar)
+
+# prior:
+
+prior_diss <- c(
+  prior(normal(0.5, .25), class = ar, lb = 0, ub = 1),
+  prior(student_t(3, 0, 2.5), class = b),
+  prior(gamma(2, .5), class = nu),
+  prior(normal(0, .5), class = sds, coef = `s(date_yday, bs = "cc")`)
+)
+
+prior_part <- prior_diss[-4,] # remove seasonal smooth

### 2022-05-05 R/loo-cv.R

@@ -1,0 +1,27 @@
+
+# approximate leave-one-out cross-validation for models (runs in ~ 10 mins):
+
+source("R/models.R", echo = TRUE)
+library("dplyr")
+
+#------------------ loo ------------------
+
+loo_diss_car <- loo_cv(model_in, model_diss)
+loo_diss_noar <- loo_cv(model_in, model_diss_noar, car1 = FALSE)
+
+loo_part_car <- loo_cv(model_in, model_part)
+loo_part_noar <- loo_cv(model_in, model_part_noar, car1 = FALSE)
+
+# CAR(1) models fit better:
+loo_compare(loo_diss_car, loo_diss_noar)
+loo_compare(loo_part_car, loo_part_noar)
+
+# write pareto-k values to file:
+model_in %>%
+  transmute(
+    location, ortho_dose, pipe_material,
+    date, date_numeric, 
+    pk_vals_diss = loo::pareto_k_influence_values(loo_diss_car),
+    pk_vals_part = loo::pareto_k_influence_values(loo_part_car)
+  ) %>%
+  readr::write_csv("data-clean/pareto-k-values.csv")

### 2022-05-05 R/models.R

@@ -1,0 +1,84 @@
+
+# fit GAMs to data and simulated data (models are loaded from CSVs by default):
+
+#------------------ setup ------------------
+
+source("R/formulas.R")
+library("rstan")
+library("bgamcar1")
+
+options(mc.cores = parallel::detectCores())
+
+#------------------ inputs ------------------
+
+model_in <- readr::read_csv("data-clean/model_in.csv")
+model_in_sim <- readr::read_csv("data-clean/simulated-data.csv")
+
+knots_yday <- c(0, 1)
+
+#------------------ models ------------------
+
+# n.b., each takes ~30 min to fit on a 2017 Macbook Pro
+
+model_diss <- fit_stan_model(
+  "models/model_diss",
+  seed = stan_seed,
+  form_diss, model_in, prior_diss,
+  save_warmup = FALSE,
+  control = list(max_treedepth = 12),
+  knots = list(date_yday = knots_yday)
+)
+
+model_part <- fit_stan_model( 
+  "models/model_part",
+  seed = stan_seed,
+  form_part, model_in, prior_part, 
+  save_warmup = FALSE
+)
+
+model_diss_noar <- fit_stan_model(
+  "models/model_diss_noar",
+  seed = stan_seed,
+  form_diss_noar, model_in, prior_diss[-1,],
+  save_warmup = FALSE,
+  control = list(max_treedepth = 12), 
+  knots = list(date_yday = knots_yday)
+)
+
+model_part_noar <- fit_stan_model(
+  "models/model_part_noar",
+  seed = stan_seed,
+  form_part_noar, model_in, prior_part[-1,],
+  save_warmup = FALSE
+)
+
+model_sim_diss <- fit_stan_model(
+  "models/model_sim_diss",
+  seed = stan_seed,
+  form_diss, model_in_sim, prior_diss,
+  save_warmup = FALSE,
+  knots = list(date_yday = knots_yday)
+)
+
+model_sim_part <- fit_stan_model(
+  "models/model_sim_part",
+  seed = stan_seed,
+  form_part_sim, model_in_sim, prior_part,
+  save_warmup = FALSE
+)
+
+model_noprior_diss <- fit_stan_model(
+  "models/model_noprior_diss",
+  seed = stan_seed,
+  form_diss, model_in, 
+  save_warmup = FALSE,
+  control = list(max_treedepth = 12), 
+  knots = list(date_yday = knots_yday) 
+)
+
[…]

### 2022-05-05 R/simulate-data.R

@@ -1,0 +1,73 @@
+
+# simulate data from the final model for fitting a test model:
+
+source("R/models.R", echo = TRUE)
+library("dplyr")
+library("purrr")
+library("withr")
+library("tidyr")
+library("readr")
+library("assertr", include.only = "verify")
+
+#------------------ simulate ------------------
+
+preds_raw <- list("lead_dissolved" = model_diss, "lead_part" = model_part) %>% 
+  map_dfr(
+    ~ withr::with_seed(432, {
+      add_pred_draws_car1(
+        model_in, .x,
+        draw_ids = 3386, type = "prediction"
+      )
+    }),
+    .id = "type"
+  )
+
+preds_raw %>%
+  ungroup() %>%
+  distinct(type, `ar[1]`, sigma, nu)
+
+preds <- preds_raw %>% 
+  pivot_wider(
+    id_cols = names(model_in),
+    names_from = type, values_from = .prediction, names_prefix = ".prediction_"
+  )
+
+cens_lim_scaled <- model_in %>%
+  filter(cens_lead_dissolved == "left", lead_dissolved == min(lead_dissolved)) %>%
+  distinct(lead_dissolved, scaled_lead_dissolved)
+
+simdat <- preds %>%
+  ungroup() %>%
+  mutate(
+    # modify censoring indicators:
+    across(
+      starts_with(".prediction_"),
+      ~ if_else(.x < cens_lim_scaled$scaled_lead_dissolved, "left", "none"), 
+      .names = "cens_{str_remove(.col, '.prediction_')}"
+    ),
+    # recensor:
+    across(
+      starts_with(".prediction_"), 
+      ~ pmax(.x, cens_lim_scaled$scaled_lead_dissolved),
+      .names = "scaled_{str_remove(.col, '.prediction_')}"
+    ),
+  )
+
+# final checks:
+
+simdat %>%
+  assertr::verify(scaled_lead_dissolved < 10 * max(model_in$scaled_lead_dissolved)) %>%
+  assertr::verify(scaled_lead_part < 10 * max(model_in$scaled_lead_part)) %>%
+  assertr::verify(length(scaled_lead_part) == length(model_in$scaled_lead_part)) %>%
+  assertr::verify(!is.na(scaled_lead_part)) %>%
+  assertr::verify(!is.na(scaled_lead_dissolved)) %>%
+  # censored values are all 2e-4, uncensored greater:
+  assertr::verify(
[…]

### 2022-05-05 R/tests/tests.R

@@ -1,0 +1,91 @@
+
+# this script includes:
+# (1) a few additional tests of the "bgamcar1" functions
+# (2) model diagnostics
+
+#------------------ tests ------------------
+
+library("testthat")
+source("R/models.R", echo = TRUE)
+library("purrr")
+library("dplyr")
+library("posterior")
+
+#------------------ tests ------------------
+
+# loo_cv():
+
+loo1 <- loo(model_diss, incl_autocor = FALSE)
+loo2 <- loo_cv(model_in, model_diss, car1 = FALSE)
+
+loo3 <- loo(model_part, incl_autocor = FALSE)
+loo4 <- loo_cv(model_in, model_part, car1 = FALSE)
+
+test_that("brms::loo() yields the same results as loo_cv()", {
+  expect_equal(loo1$estimate, loo2$estimate)
+  expect_equal(loo3$estimate, loo4$estimate)
+})
+
+# --------------- model diagnostics ---------------
+
+model_list <- list(
+  "diss" = model_diss,
+  "part" = model_part,
+  "diss_noar" = model_diss_noar,
+  "part" = model_part_noar,
+  # simulated data
+  "sim_diss" = model_sim_diss,
+  "sim_part" = model_sim_part
+)
+
+sampler_params <- model_list %>%
+  map_dfr(
+    # warmup draws should all be excluded b/c they aren't saved in the CSV files
+    ~ get_sampler_params(.x$fit, inc_warmup = FALSE) %>%
+      map(as_tibble) %>%
+      bind_rows(.id = ".chain") %>%
+      mutate(max_treedepth = .x$fit@stan_args[[4]]$max_treedepth),
+    .id = "model"
+  )
+
+draw_summary <- model_list %>%
+  map_dfr(
+    ~ as_draws(.x) %>%
+      posterior::summarise_draws(),
+    .id = "model"
+  )
+
+test_that("max_treedepth never exceeded", {
+  exceedances <- sampler_params %>%
+    filter(treedepth__ >= max_treedepth)
+  n <- nrow(exceedances)
+  expect_equal(n, 0)
+})
+
+test_that("no divergent transitions", {
+  divergences <- sampler_params %>%
+    filter(divergent__ != 0)
+  n <- nrow(divergences)
+  expect_equal(n, 0)
+})
+
+test_that("no rhats greater than 1.05", {
[…]

### 2022-05-05 README.Rmd
added readme

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Comparing corrosion control treatments using a robust Bayesian generalized additive model

This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model".

### 2022-05-05 R/check-pp.R

@@ -1,0 +1,44 @@
+
+# generate a density overlay plot with all of the draws (runs in ~ 10 mins.)
+
+#------------------ setup ------------------
+
+source("R/models.R", echo = TRUE)
+library("ggplot2")
+library("dplyr")
+
+theme_set(theme_bw() + theme(legend.position = "bottom"))
+palette <- wesanderson::wes_palette("Zissou1")
+
+#------------------ predict ------------------
+
+preds <- add_pred_draws_car1(
+  model_in, model_diss,
+  # draw_ids = withr::with_seed(12478, {sample(1:4000, 50)}),
+  type = "prediction"
+)
+
+cens_bound <- model_in %>%
+  filter(cens_lead_dissolved == "left") %>%
+  summarize(bound = min(scaled_lead_dissolved))
+
+yrep <- preds %>%
+  ungroup() %>%
+  # recensor:
+  mutate(scaled_lead_dissolved = if_else(.prediction < cens_bound$bound, cens_bound$bound, .prediction)) %>%
+  filter(scaled_lead_dissolved < 10)
+
+#------------------ plot ------------------
+
+model_in %>%
+  ggplot(aes(scaled_lead_dissolved)) +
+  geom_density(
+    data = yrep,
+    aes(group = .draw),
+    col = palette[2],
+    size = .2
+  ) +
+  geom_density() +
+  coord_cartesian(xlim = c(-2, 10)) +
+  labs(x = "Observation/prediction", y = "Density")
+

### 2022-05-05 R/figures.R

@@ -6,8 +6,8 @@
 source("R/models.R", echo = TRUE)
 library("forcats")
 library("ggplot2")
-library("dplyr") 
-library("tidyr") 
+library("dplyr")
+library("tidyr")
 library("purrr")
 library("patchwork")
 library("readr")
@@ -48,26 +48,26 @@ annotations <- tibble::tribble(
   "2020-03-01", 0.1, "3. Autocorrelation\n(clustering of similar\nvalues in time)",
   "2018-04-01", .7, "4. Irregular sampling\nfrequency",
   "2021-06-01", .5, "5. Extreme\nvalues",
-) %>% 
+) %>%
   mutate(date = as.Date(date))
 
-fig1 <- model_in %>% 
-  filter(location == "LP31") %>% 
-  pivot_longer(c(lead_part, lead_dissolved), names_to = "type") %>% 
-  ggplot(aes(date, value)) + 
-  geom_line(aes(col = type)) + 
+fig1 <- model_in %>%
+  filter(location == "LP31") %>%
+  pivot_longer(c(lead_part, lead_dissolved), names_to = "type") %>%
+  ggplot(aes(date, value)) +
+  geom_line(aes(col = type)) +
   geom_rug(sides = "t", color = "grey", length = unit(.02, "npc")) +
   geom_label(
     data = annotations,
     aes(label = label),
-    alpha = .8, label.r = unit(0, "cm"), 
+    alpha = .8, label.r = unit(0, "cm"),
     label.size = 0,
     size = 2
   ) +
   scale_y_log10(
     limits = c(1e-4, 2),
     labels = function(breaks) 1e3 * breaks
-  ) + 
+  ) +
   scale_color_manual(
     values = palette[c(5, 1)],
     labels = c("<0.45 µm", ">0.45 µm")
@@ -336,42 +336,42 @@ calc_spag_smooths <- function(model_diss, model_part) {
   expand_grid(
     model = list(model_diss, model_part),
     smooth_term = c(
-      "s(date_numeric)", 
-      "s(date_yday,bs=\"cc\")", 
+      "s(date_numeric)",
+      "s(date_yday,bs=\"cc\")",
       "s(date_numeric,by=location,m=1)"
     )
-  ) %>% 
-    mutate(resp = map(model, ~ attributes(.x$fit)$model_name)) %>% 
-    unnest(resp) %>% 
[…]

### 2022-05-05 R/check-prior.R
code comments

@@ -1,5 +1,5 @@
 
-# prior predictive checks (model is loaded from CSVs by default):
+# prior predictive checks (models are loaded from CSVs by default):
 
 #------------------ setup ------------------
 

### 2022-05-05 Rmarkdown/paper-si.Rmd
preprint submission

@@
 
-This document contains 5 figures, 4 tables, and 10 pages.
+This document contains 8 figures, 4 tables, and 12 pages.
 
@@
 
-```{r figure-s5, fig.cap="**Figure S5.** In the model fitted to simulated data, **(a)** Seasonal component of the trend in dissolved lead.
+```{r figure-s5, fig.cap="**Figure S5.** In the model fitted to simulated data, **(a)** seasonal component of the trend in dissolved lead.
 **(b)** Global multiyear component of the trend in dissolved and particulate lead.

### 2022-05-05 Rmarkdown/paper.Rmd
preprint submission

@@
 Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data.
-And while rank-based procedures such as the rank-sum test can accomodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].
+And while rank-based procedures such as the rank-sum test can accommodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].
 
@@
 The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2).
-This is because it accounts for non-linear patterns that the simpler models ignore.
+This is because it accounts for non-linear patterns that the simpler models ignore [@trueman_seasonal_2022].
 GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).
@@
 The lower bound of the interval was set as the difference between the total lead concentration and the detection limit and the upper bound as the total lead concentration.
-Given their impractically high censoring rates (up to 96%), time series representing effluent from soldered copper pipe sections were omitted.
+Given their impractically high censoring rates (up to 96%), time series representing effluent from soldered copper pipe sections were not modeled.
 
@@
 The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a).
-Both models included group-level smooths describing multiyear variation in lead release from each pipe (Figure 4c), a heirarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019].
+Both models included group-level smooths describing multiyear variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019].
 The multiyear smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
@@
 where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$.
-The autocorrelation coeffient $\phi^s$ is defined in equation $(4)$.
+The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$.
 
@@
 
-### Estimating derivatives
+### Estimating rates of change
 
@@
 Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation.
-The distribution of the autocorrelation coefficient $\phi$ is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.
+The distribution of the autocorrelation coefficient for each model, $\phi$, is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.
 
@@
 Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$.
-The effect of the autocorrelation term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe copuples in Nov.
+The effect of the autocorrelation term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov.
 2020 ("Pb-Cu", Figure 6).
@@
 The expected value of the posterior predictive distribution (*i.e.*, $\\mu_t$ in Equation 5) is summarized by the median (blue and red lines) and the 2.5--97.5^th^ percentiles (red/blue shaded regions).
-The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections, the yellow shaded bands represent brief periods when orthophosphate dosing was interrupted, and black ticks at the top of each plot represent values outside the plot limits.
+The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections, the yellow shaded bands represent brief periods when orthophosphate dosing was interrupted, and ticks at the top of each panel represent values outside the plot limits.
 Other significant disturbances are labeled, as described in the section titled *Model distribution system*"}
@@
 The ratio comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b).
-As a comparison of either particulate or lead concentrations, it was less than one over much of the first part of the series and greater than one over much of the latter part.
-The transition occurred a short time after orthophosphate was decreased from 2--0.75 mg P L^-1^, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose, likely as particles and small colloids.
+As a comparison of either particulate or dissolved lead concentrations, it was less than one over much of the first part of the series and greater than one over much of the latter part.
+The transition occurred a short time after orthophosphate was decreased from 2--0.75 mg P L^-1^, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose, likely as particles and small colloids [@trueman_characterizing_2019; @li_impact_2020].
 Meanwhile, this dose presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.
@@
 
-Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the duration of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
+Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
 This general pattern is represented in our model by the global multiyear trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature).
@@
 Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
-Both fractions exhibited more modest changes in subsequent years that are difficult to attribute, but they may be a response to changes in water quality over the course of the study.
+Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to changes in water quality over the course of the study.
 
 ```{r figure-8, fig.cap="**Figure 8.
-(a) ** The global multiyear smooth and its estimated first derivative.
-A random sample of 200 smooths/derivatives is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions.
-The median of the global trend is shaded red when the 95% credible interval around the first derivative does not include zero.
+(a) ** The global multiyear smooth and its estimated local slope.
+A random sample of 200 smooths/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions.
+The median of the global trend is shaded red when the 95% credible interval around the local slope does not include zero.
 **(b--c)** Iron and water temperature, both measured in pipe rack influent, exhibited comparable seasonal patterns to lead, the latter described by the seasonal term in the GAM.
@@
 
-Over the duration of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9).
+Over the course of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9).
 Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible.
@@
 The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the ratio difference between the 0 and 1 mg P L^-1^ dose.
-The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) yielded the lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^.
+The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) yielded lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^.
 Consistent with previous work, orthophosphate tended to increase the particulate fraction of total lead.

### 2022-05-09 README.Rmd
added DOI

@@
 
-This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model".
+This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72).
+Bespoke functions used to fit the models are available as a separate package, [`bgamcar1`](https://github.com/bentrueman/bgamcar1).

### 2022-05-09 Rmarkdown/paper-si.Rmd
added DOI

@@
 title: "Supplementary information for *Comparing corrosion control treatments using a robust Bayesian generalized additive model*"
-author: "Benjamin F. Trueman^§^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"
-date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, NS, CAN B3H 4R2    \n&nbsp;    \n&nbsp;^†^EPCOR Water Services, Edmonton, AB, CAN"
+author: "Benjamin F. Trueman^§,\\*^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"
+date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, NS, CAN B3H 4R2    \n&nbsp;    \n&nbsp;^†^EPCOR Water Services, Edmonton, AB, CAN    \n&nbsp;    \n&nbsp;^\\*^Corresponding author: benjamin.trueman@dal.ca"
 output:

### 2022-05-09 Rmarkdown/paper.Rmd
added DOI

@@
 title: "Comparing corrosion control treatments using a robust Bayesian generalized additive model"
-author: "Benjamin F. Trueman^§^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"
-date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, NS, CAN B3H 4R2    \n&nbsp;    \n&nbsp;^†^EPCOR Water Services, Edmonton, AB, CAN"
+author: "Benjamin F. Trueman^§,\\*^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"
+date: "^§^Centre for Water Resources Studies, Department of Civil & Resource Engineering, Dalhousie University, 1360 Barrington St., Halifax, NS, CAN B3H 4R2    \n&nbsp;    \n&nbsp;^†^EPCOR Water Services, Edmonton, AB, CAN    \n&nbsp;    \n&nbsp;^\\*^Corresponding author: benjamin.trueman@dal.ca"
 output:

### 2022-05-09 R/figures.R
added local_slope()

@@ -13,7 +13,6 @@ library("patchwork")
 library("readr")
 library("posterior")
 library("withr", include.only = "with_seed")
-library("imputeTS", include.only = "na_interpolation")
 library("ggdist", include.only = c("stat_halfeye", "median_qi"))
 library("glue", include.only = "glue")
 library("assertr", include.only = "verify")
@@ -901,97 +900,67 @@ grid_2 <- grid_1 %>%
 
 grid_avg <- (grid_1 + grid_2) / 2
 
-smooths <- tibble(
-  grid = list(grid_1, grid_2),
-  model = list(model_diss, model_part)
+slopes <- list(
+  "<0.45 µm" = model_diss,
+  ">0.45 µm" = model_part
 ) %>%
-  cross_df() %>%
-  mutate(
-    smooth = map2(
-      model, grid,
-      ~ posterior_smooths(.x, smooth = "s(date_numeric)", .y)
-    ),
-    start = map(grid, ~ pull(.x)[1]),
-    resp = map(model, ~ attributes(.x$fit)$model_name)
+  map_dfr(
+    ~ local_slope(grid_avg, .x, "date_numeric", smooth = "s(date_numeric)"),
+    .id = "fraction"
   ) %>%
-  unnest(c(start, resp)) %>%
-  mutate(bound = if_else(start == max(start), "upper", "lower")) %>%
-  pivot_wider(-c(model, grid, start), names_from = bound, values_from = smooth) %>%
+  rename(Smooth = smooth, "Local slope" = slope) %>%
+  pivot_longer(c(Smooth, `Local slope`))
+
+slopes_sum <- slopes %>%
+  select(-.draw) %>%
+  group_by(fraction, date_numeric, name) %>%
+  summarize_preds(pred_var = "value", retrans = FALSE)
+
+slopes_sig <- slopes_sum %>%
+  pivot_wider(
+    id_cols = c(fraction, date_numeric),
+    names_from = c(name),
+    values_from = c(value, .lower, .upper)
+  ) %>%
+  arrange(fraction, date_numeric) %>%
   mutate(
-    # grids:
-    grid_1 = list(grid_1),
-    grid_avg = list(grid_avg),
-    # summary stats:
-    smooths = map(lower, ~ apply(.x, 2, ggdist::median_qi) %>% bind_rows()),
-    # derivatives and summary stats:
-    derivs = map2(lower, upper, ~ (.y - .x) / epsilon),
[…]

### 2022-05-16 Rmarkdown/paper.Rmd
update to model notation

@@
 A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature.
-(This component may not be necessary if there is no distinct seasonality.)
+(This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it.)
 4. An autocorrelated error term.
@@
 (a)** Seasonal component of the trend in dissolved lead.
-**(b)** Global multiyear component of the trend in dissolved and particulate lead.
-**(c)** Group-level (i.e., pipe-specific) multiyear trends in dissolved and particulate lead, representing the deviations of each time series from the global multiyear trend.
+**(b)** Global longterm component of the trend in dissolved and particulate lead.
+**(c)** Group-level (i.e., pipe-specific) longterm trends in dissolved and particulate lead, representing the deviations of each time series from the global longterm trend.
 In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue.
@@
 
-The models included one-dimensional global smooths describing multiyear variation in lead across all time series (Figure 4b).
+The models included one-dimensional global smooths describing longterm variation in lead across all time series (Figure 4b).
 The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a).
-Both models included group-level smooths describing multiyear variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019].
-The multiyear smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
+Both models included group-level smooths describing longterm variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019].
+The longterm smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
 To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.
@@
 y_t | censored_t = 2 \sim G(\mu_t,\sigma,\nu) \\
-\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s(y_{t-s} - \mu_{t-s}) \\
+\mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
 \sigma \sim T(0, 2.5, 3) \\
@@
 where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$.
-The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$.
-
 Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that a random variable $X$ is at most $x$) [@stan2021; @mcelreath_statistical_2016].
@@
 $$
-
-And the $f_j(t)$ are smooth spline functions that comprise the GAM, each with the following form:
+The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$ ($r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$.
+The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smooths takes the following form:
 
@@
 
-where $Z_j$ and $X_j$ are design matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function of the global multiyear trend).
+where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global longterm trend).
 The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooths [@mgcv].
@@
 
-We estimated the local slope of the global multiyear smooths using a finite difference method.
+We estimated the local slope of the global longterm smooths using a finite difference method.
 Specifically, we generated 4000 posterior predictions of the smooth trend along a regular sequence spanning the date range of the data.
@@
 The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a).
-To estimate the time-varying mean representing each series, we modeled lead as the sum of a global multiyear smooth, a global seasonal smooth (dissolved lead only), a group-level multiyear smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).
+To estimate the time-varying mean representing each series, we modeled lead as the sum of a global longterm smooth, a global seasonal smooth (dissolved lead only), a group-level longterm smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).
 
@@
 Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
-This general pattern is represented in our model by the global multiyear trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature).
+This general pattern is represented in our model by the global longterm trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature).
 The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term.
@@
 ```{r figure-8, fig.cap="**Figure 8.
-(a) ** The global multiyear smooth and its estimated local slope.
+(a) ** The global longterm smooth and its estimated local slope.
 A random sample of 200 smooths/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions.
@@
 
-Supplementary information for this article includes additional description of the pipe loop study, a sensitivity analysis of the priors, models summaries, and convergence statistics.
+Supplementary information for this article includes additional description of the pipe loop study, a comparison of different prior choices, models summaries, and convergence statistics.
 

### 2022-05-24 README.Rmd
updated ref list

@@
 
-This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72).
+This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72-v2).
 Bespoke functions used to fit the models are available as a separate package, [`bgamcar1`](https://github.com/bentrueman/bgamcar1).

### 2022-05-24 Rmarkdown/paper.Rmd
updated ref list

@@
 
-Pipe loop studies are used routinely to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain an important tool for water quality management.
-But they are inherently complex and the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common features, the latter due to repeatedly measuring the effluent from test pipes.
-Popular statistical tests, such as the Student *t* or rank-sum tests, are often inadequate descriptions of the data.
-Here, we propose an approach to statistical analysis of pipe loop data that accommodates many of these difficult-to-model characteristics: a robust Bayesian generalized additive model with continuous-time autoregressive errors.
-Our model facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers.
-It is well-suited to describing nonlinear trends without overfitting, and it accounts for reduced information content due to autocorrelation.
-We demonstrate the model using an example pipe loop study comprising four years of data, multiple pipe configurations, and multiple orthophosphate dosing protocols.
-We compare the experimental treatments, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term.
-We also find that---consistent with previous work---galvanic corrosion yields relatively high particulate lead release, especially at higher orthophosphate doses.
-An advantage of the Bayesian approach we adopt here is that it yields the full posterior distribution of all parameters and all quantities derived from those parameters, including the model predictions.
-This means that analysts can construct any comparison that is relevant to the study goals, without the need to rely on a particular statistical test.
+Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for water quality management.
+But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common features that make popular methods, such as the *t*- or rank-sum tests, poor descriptive models.
+Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model characteristics: a robust Bayesian generalized additive model with continuous-time autoregressive errors.
+Our approach facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers.
+It is well-suited to describing nonlinear trends without overfitting, and it accounts for the reduced information content in autocorrelated time series.
+We demonstrate it using a four-year pipe loop study, with multiple pipe configurations and orthophosphate dosing protocols, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term.
 
@@
 Pipe loop studies are used to model drinking water distribution systems, mainly for evaluating corrosion control treatment.
-The US EPA's Revised Lead and Copper Rule is expected to result in many water utilities conducting pipe loop studies [@estessmargiassi_understanding_2020], and the data they generate will need to be analyzed.
+The US EPA's Revised Lead and Copper Rule mandates them under certain conditions [@estessmargiassi_understanding_2020; @us2021national], and the data they generate will need to be analyzed.
 But pipe loop study data are inherently complex and there is a real risk of misinterpreting them.
@@
 
-All of the above present barriers to reliable statistical inference, but left-censoring and autocorrelation can be particularly challenging.
+Left-censoring and autocorrelation can be particularly challenging to model.
 Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values.
@@
 1. A global---possibly nonlinear---trend describing variation in the parameter of interest over time across all pipes.
-2. A set of group-level trends that describe the deviations from the global trend particular to the individual pipes.
+2. A set of group-level trends that describe the deviations from the global trend particular to each individual pipe.
 3.
@@
 (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it.)
-4. An autocorrelated error term.
+4. An error term, possibly autocorrelated.
 
-We demonstrate this approach using an example pipe loop dataset representing four years of data, multiple pipe configurations, and multiple orthophosphate dosing protocols.
+We demonstrate this approach using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols.
 Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.
 An advantage of our Bayesian approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values.
-This means that we can construct any comparison we wish to investigate, without the need to decide on a particular statistical test.
+This allows for full propagation of uncertainty through all stages of post-processing and analysis [@beck_multi-scale_2022].
 Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in lead concentrations over time.
@@
 Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively.
-An aliquot of each sample was filtered at 0.45 µm using a teflon filter and a vacuum-filter apparatus.
+An aliquot of each sample was filtered at 0.45 µm using a teflon filter and a vacuum filter apparatus.
 Temperature was measured using an infrared thermometer, total chlorine [@sm_4500_cl_g] and orthophosphate [@sm_4500_p_e] were measured using a spectrophotometer, and pH was measured using a benchtop meter.
@@
 The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2).
-This is because it accounts for non-linear patterns that the simpler models ignore [@trueman_seasonal_2022].
+This is because it accounts for non-linear patterns that the simpler models ignore [@trueman_seasonal_2022; @murphy_generalized_2019].
 GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).
@@
 
-$$(4)~~h(s, \phi) = \phi^s, s\geq0, \phi\geq0, \phi\leq1 $$
+$$(4)~~h(s, \phi) = \phi^s, s\geq0, 0 \leq \phi \leq1 $$
 
@@
 We generated a code template using the R package `brms` [@brms; @brms_adv] and modified it to specify the continuous-time autocorrelation structure defined in equation $(4)$.
-We used a suite of common R packages [@tidyverse; @nada; @assertr; @janitor; @datatable], and bespoke functions used to fit the model and process the results are documented on GitHub [@bgamcar1].
+We used a suite of R packages [@tidyverse; @nada; @assertr; @janitor; @datatable; @patchwork; @posterior; @ggdist; @wesanderson; @ggh4x; @broom; @ggtext], and bespoke functions used to fit the model and process the results are documented on GitHub [@bgamcar1].
 The data and R code necessary to reproduce the analysis are available in a separate repository [@compare_cct].
@@
 (a)** Seasonal component of the trend in dissolved lead.
-**(b)** Global longterm component of the trend in dissolved and particulate lead.
-**(c)** Group-level (i.e., pipe-specific) longterm trends in dissolved and particulate lead, representing the deviations of each time series from the global longterm trend.
+**(b)** Global long-term component of the trend in dissolved and particulate lead.
+**(c)** Group-level (i.e., pipe-specific) long-term trends in dissolved and particulate lead, representing the deviations of each time series from the global long-term trend.
 In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue.
@@
 
-The models included one-dimensional global smooths describing longterm variation in lead across all time series (Figure 4b).
+The models included one-dimensional global smooths describing long-term variation in lead across all time series (Figure 4b).
 The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a).
-Both models included group-level smooths describing longterm variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019].
-The longterm smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
+Both models included group-level smooths describing long-term variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019].
+The long-term smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
 To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.
@@
 $$
-(5)
+(5)~~
 \begin{align}
@@
 $$
-(6)
+(6)~~
 \begin{align}
@@
 $$
-The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$ ($r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$.
+The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$.
 The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smooths takes the following form:
@@
 $$
-(7)
+(7)~~
 \begin{align}
@@
 
-where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global longterm trend).
+where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global long-term trend).
 The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooths [@mgcv].
@@
 
-We estimated the local slope of the global longterm smooths using a finite difference method.
+We estimated the local slope of the global long-term smooths using a finite difference method.
 Specifically, we generated 4000 posterior predictions of the smooth trend along a regular sequence spanning the date range of the data.
@@
 
-# Results
+# Results and discussion
 
@@
 The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a).
-To estimate the time-varying mean representing each series, we modeled lead as the sum of a global longterm smooth, a global seasonal smooth (dissolved lead only), a group-level longterm smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).
+To estimate the time-varying mean representing each series, we modeled lead as the sum of a global long-term smooth, a global seasonal smooth (dissolved lead only), a group-level long-term smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).
 
@@
 Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
-This general pattern is represented in our model by the global longterm trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature).
+This general pattern is represented in our model by the global long-term trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature).
 The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term.
@@
 ```{r figure-8, fig.cap="**Figure 8.
-(a) ** The global longterm smooth and its estimated local slope.
+(a) ** The global long-term smooth and its estimated local slope.
 A random sample of 200 smooths/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions.
@@
 Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible.
-The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018].
+The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017].
 

### 2022-05-25 Rmarkdown/paper.Rmd
new refs, updated figs

@@
 ---
+
+<!-- make sure to redo figure s5 in SI -->
 
@@
 
-The continuous-time autocorrelated error term was successful in capturing residual first-order autocorrelation, estimated in Figure 5b using the Spearman rank correlation coefficient as a measure that is robust against outliers.
-Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset.
-Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation.
-The distribution of the autocorrelation coefficient for each model, $\phi$, is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.
-
 ```{r figure-5, fig.cap="**Figure 5.
-(a) ** Histograms representing the median of 4000 residual draws from the models; the theoretical densities are calculated from the posterior medians of the distributional parameters (i.e., $\\sigma$, $\\nu$) from each model.
-**(b)** Density of first-order autocorrelation, estimated using the Spearman rank correlation coefficient and 4000 residual draws from the model including and excluding the continuous-time first-order autocorrelation term.
+(a) ** Histograms representing the median of 4000 residual draws from the models; the theoretical densities (red curves) are calculated from the posterior medians of the distributional parameters (i.e., $\\sigma$, $\\nu$) from each model.
+**(b)** Density of first-order autocorrelation, estimated using the Spearman rank correlation coefficient and 4000 residual draws from the model including and excluding the continuous-time first-order autocorrelation term (GAM and CAR(1), respectively).
 Estimates were generated using an augmented dataset where censored observations were imputed via posterior prediction.
@@
 ```
+
+The continuous-time autocorrelated error term was successful in capturing residual first-order autocorrelation, estimated in Figure 5b using the Spearman rank correlation coefficient as a measure that is robust against outliers.
+Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset.
+Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation.
+The distribution of the autocorrelation coefficient for each model, $\phi$, is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.
 
@@
 Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$.
-The effect of the autocorrelation term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov.
+The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov.
 2020 ("Pb-Cu", Figure 6).
@@
 Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
-Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to changes in water quality over the course of the study.
+Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to changes in water quality over the study.
 
@@
 
+```{r figure-9, fig.cap="**Figure 9.** The particulate fraction of total lead over time, grouped by pipe and orthophosphate dosing protocol, and estimated from the posterior distributions of the models fitted to particulate and dissolved lead time series.
+The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections."}
+knitr::include_graphics(here::here("Rmarkdown/figures/figure-9.png"))
+```
+
 ## Estimating the particulate fraction
 
-Over the course of the study, a majority of total lead was present in 0.45 µm filtrate (Figure 9).
+Across all pipe sections, a majority of total lead was present in 0.45 µm filtrate (Figure 9).
 Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible.
-The particulate fraction was also greater in the effluent from pipes featuring a galvanic junction, and both patterns are consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017].
-
-```{r figure-9, fig.cap="**Figure 9.** The particulate fraction of total lead over time, grouped by pipe and orthophosphate dosing protocol, and estimated from the posterior distributions of the models fitted to particulate and dissolved lead time series."}
-knitr::include_graphics(here::here("Rmarkdown/figures/figure-9.png"))
-```
+During the conditioning period---before orthophosphate was introduced---the particulate fraction depended mainly on the pipe configuration.
+Almost immediately after conditioning, the particulate fraction increased in the first two arrays as orthophosphate decreased lead solubility.
+The inverse seasonality---higher particulate fractions in winter---is a function of the relatively constant particulate lead release and the strongly seasonal dissolved lead release (Figure 6).
+The particulate fraction was also higher in the effluent from pipes featuring a galvanic junction.
+This, and the effect of orthophosphate on lead speciation, is consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
 
@@
 Moreover, the model accounts for the reduced information content in autocorrelated observations.
-We demonstrate it on an example pipe loop dataset and use the model to compare orthophosphate dosing protocols and analyze time trends.
+We demonstrate it using data from a four-year pipe loop study and we use the model to compare orthophosphate dosing protocols and analyze time trends.
 The additive smooth terms comprising the model yield an informative decomposition of pipe loop time series and a more nuanced comparison of corrosion control treatments.
@@
 The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the ratio difference between the 0 and 1 mg P L^-1^ dose.
-The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) yielded lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^.
+The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) often yielded lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^.
 Consistent with previous work, orthophosphate tended to increase the particulate fraction of total lead.

### 2022-05-25 R/figures.R

@@ -42,11 +42,11 @@ locations <- model_in %>%
 
 annotations <- tibble::tribble(
   ~date, ~value, ~label,
-  "2019-01-01", 0.02, "1. Non-linearity",
+  "2018-12-01", 0.01, "1. Non-linearity",
   "2020-10-01", 1e-4, "2. Left-censoring",
-  "2020-03-01", 0.1, "3. Autocorrelation\n(clustering of similar\nvalues in time)",
+  "2020-03-01", 0.04, "3. Autocorrelation\n(clustering of similar\nvalues in time)",
   "2018-04-01", .7, "4. Irregular sampling\nfrequency",
-  "2021-06-01", .5, "5. Extreme\nvalues",
+  "2021-05-01", .5, "5. Extreme\nvalues",
 ) %>%
   mutate(date = as.Date(date))
 
@@ -54,7 +54,6 @@ fig1 <- model_in %>%
   filter(location == "LP31") %>%
   pivot_longer(c(lead_part, lead_dissolved), names_to = "type") %>%
   ggplot(aes(date, value)) +
-  geom_line(aes(col = type)) +
   geom_rug(sides = "t", color = "grey", length = unit(.02, "npc")) +
   geom_label(
     data = annotations,
@@ -63,6 +62,7 @@ fig1 <- model_in %>%
     label.size = 0,
     size = 2
   ) +
+  geom_line(aes(col = type)) +
   scale_y_log10(
     limits = c(1e-4, 2),
     labels = function(breaks) 1e3 * breaks
@@ -288,7 +288,7 @@ fig3 <- gam_basis %>%
     )
   ) %>%
   ggplot(aes(date)) +
-  facet_wrap(vars(name), scales = "free_y", ncol = 2) +
+  facet_wrap(vars(name), scales = "free_y", ncol = 1) +
   geom_line(aes(y = basis_value, col = basis_name)) +
   geom_line(
     data = function(x) {
@@ -318,9 +318,10 @@ fig3 <- gam_basis %>%
     x = NULL,
     y = expression("log([Pb] (µg L"^-1 * ")"),
     col = NULL
-  )
+  ) +
+  guides(col = guide_legend(nrow = 4))
 
-ggsave("Rmarkdown/figures/figure-3.png", fig3, width = 7, height = 2.5, dpi = 600)
+ggsave("Rmarkdown/figures/figure-3.png", fig3, width = 3.33, height = 4, dpi = 600)
 
[…]

### 2022-05-30 Rmarkdown/paper-si.Rmd
journal submission

@@
 ---
-title: "Supplementary information for *Comparing corrosion control treatments using a robust Bayesian generalized additive model*"
+title: "Supplementary information for *Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model*"
 author: "Benjamin F. Trueman^§,\\*^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"

### 2022-05-30 Rmarkdown/paper.Rmd
journal submission

@@
 ---
-title: "Comparing corrosion control treatments using a robust Bayesian generalized additive model"
+title: "Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model"
 author: "Benjamin F. Trueman^§,\\*^, Wendell James^†^, Trevor Shu^†^, Evelyne Dor&eacute;^§^, and Graham A. Gagnon^§^"
@@
 ---
-
-<!-- make sure to redo figure s5 in SI -->
 
@@
 
-# Methods
+# Materials and methods
 
@@
 All plumbing was new at the time of installation, and lead pipes were acquired from Canada Metals Eastern (ASTM B29, 99.94% pure lead).
-Each array was preceded in the flow-path by a 3.6 m, 3" ID ductile iron pipe, and system pressure was approximately 40 psi.
+Each array was preceded in the flow path by a 3.6 m, 3" ID ductile iron pipe, and system pressure was approximately 40 psi.
 
@@
 2017 to Mar.
-2018 using treated water from the E.L.
+2018 using treated water from the E.
+L.
 Smith Water Treatment Plant in Edmonton, AB, CAN.
@@
 Lack of independence is apparent in Figure 2b: the residual of either model at time *t*, $\epsilon_t$, is correlated with the residual at time *t-1*, $\epsilon_{t-1}$.
-This concept is extended to include the residual at time *t-k*, $\epsilon_{t-k}$, where *k = 1...21* to yield the autocorrelation function (Figure 2c).
+This can be extended to include the residual at time *t-k*, $\epsilon_{t-k}$, where *k = 1...21*, yielding the autocorrelation function (Figure 2c).
 Independent residuals yield autocorrelation functions that hover around zero, whereas correlated residuals typically yield autocorrelation functions that either decay or oscillate.
@@
 
-$$(4)~~h(s, \phi) = \phi^s, s\geq0, 0 \leq \phi \leq1 $$
+$$(4)~~h(s, \phi) = \phi^s,~s\geq0,~0 \leq \phi \leq1 $$
 
@@
 Particulate lead was determined as the difference between total and dissolved lead.
-Particulate lead concentrations less than the MDL of 0.2 µg L^-1^ were left-censored, and particulate lead concentrations representing the difference between an observed (total) and left-censored (dissolved) concentration were interval-censored.
+Particulate lead concentrations less than the detection limit of 0.2 µg L^-1^ were left-censored, and particulate lead concentrations representing the difference between an observed (total) and left-censored (dissolved) concentration were interval-censored.
 The lower bound of the interval was set as the difference between the total lead concentration and the detection limit and the upper bound as the total lead concentration.
@@
 The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections, the yellow shaded bands represent brief periods when orthophosphate dosing was interrupted, and ticks at the top of each panel represent values outside the plot limits.
-Other significant disturbances are labeled, as described in the section titled *Model distribution system*"}
+Other significant disturbances are labeled, as described in the methods section (see *Model distribution system*)."}
 knitr::include_graphics(here::here("Rmarkdown/figures/figure-6.png"))
@@
 
-This work was supported by Mitacs, through the Mitacs Accelerate Program (Reference # IT23352), and NSERC, through an NSERC/Halifax Water Industrial Research Chair program (Grant # IRCPJ: 349838-16).
+This work was supported by Mitacs, through the Mitacs Accelerate Program (Reference # IT23352), and NSERC, through an Industrial Research Chair program (Grant # IRCPJ: 349838-16) and a Postdoctoral Fellowship (E.
+Dor&eacute;).
 We thank David Redden and Paul Bjorndahl for providing comments on a draft of the manuscript.

### 2022-05-30 R/check-prior.R

@@ -24,7 +24,7 @@ knots_yday <- c(0, 1)
 # prior simulation:
 
 model_prior_part <- fit_stan_model(
-  "data-clean/prior_part",
+  "models/prior_part",
   seed = stan_seed,
   form_part, model_in, prior_part,
   save_warmup = FALSE,
@@ -32,7 +32,7 @@ model_prior_part <- fit_stan_model(
 )
 
 model_prior_diss <- fit_stan_model(
-  "data-clean/prior_diss",
+  "models/prior_diss",
   seed = stan_seed,
   form_diss, model_in, prior_diss,
   save_warmup = FALSE,

### 2022-05-30 R/figures.R

@@ -427,7 +427,6 @@ plot_fig4 <- function(smooth_terms_diss, smooth_terms_part, smooth_terms_spag) {
     patchwork::plot_annotation(tag_level = "a") &
     theme(
       plot.tag = element_text(face = "bold"),
-      # axis.text.x = element_text(angle = 35, hjust = 1),
       legend.position = "bottom"
     ) &
     scale_color_manual(values = palette[c(5,3,1)]) &
@@ -584,7 +583,6 @@ p4 <- draws_pars %>%
 
 fig5 <- patchwork::wrap_plots(
   p2, p1, p4, p3,
-  # design = "AB\nCD"
   nrow = 4,
   heights = c(.8,1.2,.8,.8)
 ) +
@@ -1088,13 +1086,8 @@ fig8c <- pdat %>%
   scale_color_manual(values = palette[c(5,3,1)]) +
   theme(
     legend.margin = margin(),
-    # legend.background = element_rect(fill = alpha("white", .5)),
-    strip.text = ggtext::element_markdown(),
-    # legend.position = c(.5, .15),
-    # legend.direction = "horizontal"
-  ) +
-  # guides(col = guide_legend(nrow = 2)) +
-  # ylim(c(0, 27)) +
+    strip.text = ggtext::element_markdown()
+  ) +s
   labs(
     x = NULL,
     y = NULL,
@@ -1137,7 +1130,6 @@ preds_fdiss_summ <- preds_fdiss %>%
   summarize_preds(retrans = FALSE, pred_var = "fdiss")
 
 fig9 <- preds_fdiss_summ %>%
-  # mutate(ortho_dose = paste0(ortho_dose, " mg P L<sup>-1</sup>")) %>%
   ggplot(aes(date, fdiss, col = ortho_dose, fill = ortho_dose)) +
   facet_wrap(vars(pipe_material), ncol = 1) +
   geom_rect(

### 2022-05-31 README.Rmd
readme update

--- 2022-05-09 README.Rmd
@@
 
-# Comparing corrosion control treatments using a robust Bayesian generalized additive model
+# Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model
 
-This repository contains the R code and other materials for the paper "Comparing corrosion control treatments using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72).
+This repository contains the R code and other materials for the paper "Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72-v3).
 Bespoke functions used to fit the models are available as a separate package, [`bgamcar1`](https://github.com/bentrueman/bgamcar1).

### 2022-07-18 Rmarkdown/paper.Rmd
edit fig 6, add SI description, keywords

@@
 
-```{r toc-art}
-knitr::include_graphics(here::here("Rmarkdown/figures/figure-toc.png"))
-```
+## Keywords
+
+GAM; time series analysis; censored data; pipe loop; Lead and Copper Rule
 
@@
 We thank David Redden and Paul Bjorndahl for providing comments on a draft of the manuscript.
-We also acknowledge EPCOR's analytical operations team for water quality analysis and their
-process development team for operating the pipe loops.
+We also acknowledge EPCOR's analytical operations team for water quality analysis and their process development team for operating the pipe loops.
 
@@
 
-Supplementary information for this article includes additional description of the pipe loop study, a comparison of different prior choices, models summaries, and convergence statistics.
+## Supporting information
+
+The Supporting Information is available free of charge via the internet at http://pubs.acs.org.
+
+Additional description of the pipe loop study and the statistical model, including model summaries, convergence statistics, and a comparison of different prior choices.
 
 # References
+
+<div id="refs"></div>
+
+# TOC graphic
+
+```{r toc-art}
+knitr::include_graphics(here::here("Rmarkdown/figures/figure-toc.png"))
+```

### 2022-07-18 R/figures.R

@@ -27,7 +27,7 @@ library("ggtext", include.only = c("element_markdown", "geom_richtext"))
 library("tibble", include.only = c("tribble", "rowid_to_column"))
 
 theme_set(theme_bw() + theme(legend.position = "bottom"))
-palette <- wesanderson::wes_palette("Zissou1")
+palette <- wesanderson::wes_palette("Zissou1", 6, "continuous")
 
 #------------------ load ------------------
 
@@ -68,7 +68,7 @@ fig1 <- model_in %>%
     labels = function(breaks) 1e3 * breaks
   ) +
   scale_color_manual(
-    values = palette[c(5, 1)],
+    values = palette[c(6, 1)],
     labels = c("<0.45 µm", ">0.45 µm")
   ) +
   labs(x = NULL, y = expression("[Pb] (µg L"^-1*")"), col = NULL)
@@ -198,7 +198,7 @@ p2 <- models_initial %>%
     },
     aes(x = Inf, y = -Inf, label = txt),
     hjust = "inward", vjust = "inward",
-    label.size = 0, col = palette[5],
+    label.size = 0, col = palette[6],
     parse = TRUE
   ) +
   labs(
@@ -240,7 +240,7 @@ p3 <- models_initial %>%
     aes(col = "italic(r)~from~panel~bold(b)"),
     size = 2
   ) +
-  scale_color_manual(values = palette[5], labels = function(breaks) parse(text = breaks)) +
+  scale_color_manual(values = palette[6], labels = function(breaks) parse(text = breaks)) +
   scale_x_continuous(labels = function(breaks) paste0("k = ", breaks)) +
   labs(
     x = expression(italic(epsilon["t-k"])),
@@ -429,8 +429,8 @@ plot_fig4 <- function(smooth_terms_diss, smooth_terms_part, smooth_terms_spag) {
       plot.tag = element_text(face = "bold"),
       legend.position = "bottom"
     ) &
-    scale_color_manual(values = palette[c(5,3,1)]) &
-    scale_fill_manual(values = palette[c(5,3,1)])
+    scale_color_manual(values = palette[c(6,4,1)]) &
+    scale_fill_manual(values = palette[c(6,4,1)])
 }
 
[…]

### 2022-07-19 Rmarkdown/paper-si.Rmd
PLC specs, figure annotations, edits to intro

@@
   rename_all(str_to_sentence) %>% 
-  knitr::kable(caption = "**Table S3.** Pipe array flow schedule; sampling occurred at T4, and FY-001, -002, and -003 are defined in Figure S1.")
+  knitr::kable(caption = "**Table S3.** Pipe array flow schedule; sampling occurred at T4, and FY-001, -002, and -003 are defined in Figure S1.
+A Modicon M221 logic controller (Schneider Electric, Canada) was used for process control.
+Three 2-inch ASCO 8210G100 solenoid valves (i.e., FY-001, -002 and -003) were wired to the logic controller, opening and closing according to the schedule.")
 

### 2022-07-19 Rmarkdown/paper.Rmd
PLC specs, figure annotations, edits to intro

@@
 For example, the effect of a perturbation to corrosion scale is evident for some time after the fact.
-From a statistical perspective, autocorrelation makes parameter estimates more uncertain [@helsel1992statistical] by reducing the effective sample size (i.e., the information content of the sample) [@gelman_bayesian_2014].
+From a statistical perspective, autocorrelation makes parameter estimates more uncertain [@helsel_statistical] by reducing the effective sample size (i.e., the information content of the sample) [@gelman_bayesian_2014].
 It is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time.
@@
 Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014].
-As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel1992statistical].
-The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel1992statistical].
+As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel_statistical].
+The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel_statistical].
 That is, *n* autocorrelated observations provide less information than *n* independent observations, and the difference can be large if autocorrelation is strong [@gelman_bayesian_2014].
 Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data.
-And while rank-based procedures such as the rank-sum test can accommodate left-censoring, most are not designed for autocorrelated data [@helsel1992statistical; @helsel_statistics_2012].
+And while rank-based procedures such as the rank-sum test can accommodate left-censoring, most are not designed for autocorrelated data [@helsel_statistical; @helsel_statistics_2012].
+Likewise, basic parametric methods cannot describe many pipe loop datasets adequately: the paired t-test assumes random sampling of the measurement pairs [@illowsky_introductory_2018; @helsel_statistical] and simple repeated measures ANOVA has a limited capacity to model autocorrelation [@raadt_pressing_2019].
+Neither test accommodates censoring and both assume normality.
 
@@
 
-Flow through the pipe sections was programmed to approximate typical use and is summarized in Table S3.
+Flow through the pipe sections was programmed to approximate typical use, and details are available in Table S3.
 Samples were collected weekly until Mar.

### 2022-07-19 R/figures.R

@@ -41,13 +41,13 @@ locations <- model_in %>%
 #------------------ figure 1 ------------------
 
 annotations <- tibble::tribble(
-  ~date, ~value, ~label,
-  "2018-12-01", 0.01, "1. Non-linearity",
-  "2020-10-01", 1e-4, "2. Left-censoring",
-  "2020-03-01", 0.04, "3. Autocorrelation\n(clustering of similar\nvalues in time)",
-  "2018-04-01", .7, "4. Irregular sampling\nfrequency",
-  "2021-05-01", .5, "5. Extreme\nvalues",
-) %>%
+         ~date, ~value,                                                        ~label,
+  "2018-12-01",   0.01,                                            "1. Non-linearity",
+  "2020-10-01",  1e-04,                                           "2. Left-censoring",
+  "2020-03-01",   0.04, "3. Autocorrelation\n(clustering of similar\nvalues in time)",
+  "2018-04-01",    0.7,                            "4. Irregular sampling\nfrequency",
+  "2021-05-01",    0.5,                                          "5. Extreme\nvalues"
+  ) %>%
   mutate(date = as.Date(date))
 
 fig1 <- model_in %>%
@@ -598,74 +598,7 @@ ggsave("Rmarkdown/figures/figure-5.png", fig5, width = 3.33, height = 6, dpi = 6
 
 #------------------ figure 6 ------------------
 
-# n.b., dates reflect label positions;
-# correct dates are used for the vertical dashed lines
-
-annotations <- tibble::tribble(
-  ~ortho_dose, ~x, ~pipe_material, ~labels,
-  # "0-0.5", "2018-03-15", "Pb #1", "P introduced",
-  # "2.0-0.75", "2018-06-01", "Pb #1", "P decreased",
-  "0-0.5", "2020-01-01", "Pb-Cu", "Section<br>removed",
-  "1.0", "2020-01-01", "Pb-Cu", "Section<br>removed",
-  "2.0-0.75", "2020-01-01", "Pb-Cu", "Section<br>removed",
-  "0-0.5", "2017-11-01", "Pb #1", "0",
-  "0-0.5", "2017-11-01", "Pb #2", "0",
-  "0-0.5", "2017-11-01", "Pb-Cu", "0",
[…]

### 2022-07-20 Rmarkdown/paper.Rmd
new references, model description, edits to intro

@@
 But pipe loop study data are inherently complex and there is a real risk of misinterpreting them.
-The following are just some of the challenges that arise (Figure 1):
-
-(1) Non-linear time trends and seasonal variation.
-These features can be difficult to describe without over- or under-fitting.
-They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995].
-(2) Left-censored values (measurements below the detection limit), which many popular statistical tests are not designed to handle.
-(3) Autocorrelation, or memory effects.
-Autocorrelation means the current state of the system depends on its history.
-For example, the effect of a perturbation to corrosion scale is evident for some time after the fact.
-From a statistical perspective, autocorrelation makes parameter estimates more uncertain [@helsel_statistical] by reducing the effective sample size (i.e., the information content of the sample) [@gelman_bayesian_2014].
-It is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time.
-(4) An irregular sampling frequency, which makes it more difficult to estimate the degree of autocorrelation.
-(5) Extreme lead or copper concentrations due to particle release. Extreme values are often inconsistent with the common assumption of  Gaussian residuals.
+In particular, (1) non-linearity, (2) censoring, (3) autocorrelation, (4) irregular sampling frequency, and (5) outliers are difficult to model using standard approaches (Figure 1).
 
@@
 
-Left-censoring and autocorrelation can be particularly challenging to model.
+## Non-linearity
+
+Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting.
+They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995].
+The most common methods for describing pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].
+
+## Censoring
+
+Censoring typically arises in pipe loop studies when analyte concentrations fall below below the detection limit (i.e., left-censoring).
 Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values.
-Neither approach is advisable because both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012].
+Neither approach is advisable as both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012].
+Popular statistical tests (e.g., the paired t-test and its generalization, the one-way repeated measures ANOVA) cannot accommodate censoring [@helsel_statistics_2012].
+The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring).
+
+## Autocorrelation
+
+Autocorrelation means the current state of the system depends on its history.
+For example, the effect of a perturbation to corrosion scale is evident for some time after the fact.
+From a statistical perspective, autocorrelation makes parameter estimates more uncertain [@helsel_statistical] by reducing the effective sample size (i.e., the information content of the sample) [@gelman_bayesian_2014].
+It is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time, but it is not discussed in popular guides for pipe loop data analysis [@wysock_statistical_1995; @kirmeyer1996corr].
+Moreover, neither the t-test, its generalizations, nor its non-parametric analogues account for autocorrelation explicitly [@helsel_statistical; @raadt_pressing_2019].
+
 Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014].
@@
 Accounting for autocorrelation can improve predictive accuracy, but if it is ignored, analysts run the risk of making incorrect decisions based on the data.
-And while rank-based procedures such as the rank-sum test can accommodate left-censoring, most are not designed for autocorrelated data [@helsel_statistical; @helsel_statistics_2012].
-Likewise, basic parametric methods cannot describe many pipe loop datasets adequately: the paired t-test assumes random sampling of the measurement pairs [@illowsky_introductory_2018; @helsel_statistical] and simple repeated measures ANOVA has a limited capacity to model autocorrelation [@raadt_pressing_2019].
-Neither test accommodates censoring and both assume normality.
 
-Here, we propose a general approach for building statistical models of pipe loop study data that addresses all of these challenges.
-The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of four components:
+## Irregular sampling frequency
 
-1. A global---possibly nonlinear---trend describing variation in the parameter of interest over time across all pipes.
-2. A set of group-level trends that describe the deviations from the global trend particular to each individual pipe.
-3.
-A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature.
-(This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it.)
-4. An error term, possibly autocorrelated.
+An irregular sampling frequency is common in pipe loop studies where various process interruptions preclude collecting a regular measurement series.
+Irregular sampling makes it more difficult to estimate the strength and structure of autocorrelation [@simpson_modelling_2018].
 
-We demonstrate this approach using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols.
+## Outliers
+
+Extreme lead or copper concentrations---typically due to particle release---are often inconsistent with the normality assumption common to many popular statistical tests.
+Outliers are sometimes omitted [@schock_causes_1990], but unless they represent measurement error or a separate population (e.g., a process error), it is usually preferable to use a statistical method that is robust against them [@helsel_statistical].
+
+## A Bayesian approach
+
+The data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given their fairly rigid assumptions.
+Often, a more flexible approach is required, and Bayesian methods are inherently flexible, since there is no need to derive the sampling distribution of an estimator [@kruschke_doing_2015].
+
+Here, we propose a Bayesian approach for building statistical models of pipe loop study data that is sufficiently flexible for most scenarios.
+The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components:
+
+1. An average representing each pipe
+2. A global trend describing long-term variation in the parameter of interest across all pipes
+3. A set of pipe-specific trends that describe the deviations from the global long-term trend particular to each individual pipe
+4.
+A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it)
+5. An error autocorrelated term
+
+Separation of the time-trend into long-term and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the long-term trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are longer data gaps the series representing one or more pipes.
+In these instances, the model "borrows" information from the global term.
+The autocorrelated error model improves the accuracy of parameter distributions and helps to prevent overfitting.
+
+We demonstrate our approach using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols.
 Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.
@@
 Smith Water Treatment Plant in Edmonton, AB, CAN.
-Afterward, each array received a different dose of phosphoric acid: 0 mg P L^-1^ (increasing to 0.5 mg L^-1^ in Jan.
+Afterward, each array received a different dose of phosphoric acid (Table S1): 0 mg P L^-1^ (increasing to 0.5 mg L^-1^ in Jan.
 2019), 1.0 mg L^-1^, and 2.0 mg L^-1^ (decreasing to 0.75 mg L^-1^ in Apr.
 2019).
-The orthophosphate dosing schedule is summarized in Table S1.
+The latter---a high "passivation" dose followed by a lower "maintenance" dose---is discussed elsewhere as a potential corrosion control strategy [@schock1998lead].
+
 Beginning in Nov.
-2020, influent to the model system was partially dechlorinated by blending granular activated carbon filtrate with chlorinated water upstream of the pipe sections; the resulting total chlorine concentrations are summarized in Table S2.
+2020, influent to the model system was partially dechlorinated by blending granular activated carbon filtrate with chlorinated water upstream of the pipe sections.
+This was done to facilitate a biofilm study not discussed here, and the resulting total chlorine concentrations are summarized in Table S2.
 In November 2020, 30 cm sections of lead pipe were cut from the galvanic lead-copper pipes for supplementary analysis.
@@
 $$(1)~~y = \beta_{0} + f(x_t) + \epsilon_t$$ 
-where $f(x_t)$ is a smooth function of $x_t$,
+where $f(x_t)$ is a smooth function (i.e., a *smooth*) of $x_t$,
 
@@
 The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a).
-Both models included group-level smooths describing long-term variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed in a recent paper [@pedersen_hierarchical_2019].
+Both models included group-level smooths describing long-term variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019].
 The long-term smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
@@
 These can severely bias parameter estimates unless they are accounted for.
-And since our dataset was impacted by extreme lead release events, we used a Student *t* likelihood in place of the usual Gaussian likelihood to limit the influence of outliers on the posterior distribution [@brms; @brms_adv].
+And since our dataset was impacted by extreme lead release events, we used a Student *t* likelihood in place of the usual Gaussian likelihood to limit the influence of outliers [@brms; @brms_adv].
 
-The transformed data, $y_t$, were modeled as follows,
+Applying Bayes' theorem [@mcelreath_statistical_2016], the joint probability distribution of the model parameters, conditional on the data (i.e., the *posterior*), is proportional to product of the likelihood---the probability of the data, conditional on the model parameters---and the prior (i.e., the probability distribution of the model parameters uninformed by the data).
+
+<!-- $$\begin{align} -->
+<!-- P(\theta|D) \propto P(D|\theta)P(\theta) \\ -->
+<!-- posterior \propto likelihood \times prior \\ -->
+<!-- posterior = P(\theta | D) \\ -->
+<!-- likelihood = P(D | \theta) \\ -->
+<!-- prior = P(\theta) -->
+<!-- \end{align}$$ -->
+
+The transformed data, then, ($y_t$) were modeled as follows:
 
@@
 \begin{align}
+likelihood: \\
 y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu) \\ 
@@
 y_t | censored_t = 2 \sim G(\mu_t,\sigma,\nu) \\
+\\
+model: \\
 \mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
+priors: \\
 \sigma \sim T(0, 2.5, 3) \\
@@
 where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$.
-Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that a random variable $X$ is at most $x$) [@stan2021; @mcelreath_statistical_2016].
-Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$.
-The local intercepts $\alpha_{pipe_i}$ are defined as follows:
+
+Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $x$, *at most*) [@stan2021; @mcelreath_statistical_2016].
+Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$, where $x$ and $y$ are the lower and upper limits, respectively.
+Neither likelihood function requires imputation: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.
+
+The local intercepts $\alpha_{pipe_i}$ (i.e., the pipe-specific average response) are defined as follows:
 
@@
 \begin{align}
-\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha) \\
-\alpha \sim T(0.1, 2.5, 3) \\
-\sigma_{\alpha} \sim T(0, 2.5, 3) \\ 
+\alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
+\alpha \sim T(0.1, 2.5, 3)~~[prior~on~average~pipe] \\
+\sigma_{\alpha} \sim T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
 \end{align}
@@
 \begin{align}
+generalized~additive~model~(GAM): \\
 f_j(t) = X_j\beta_j + Z_jb_j \\
-\beta_j \sim T(0, 2.5, 3) \\
-b_j \sim N(0, \sigma_b) \\
+priors~for~GAM~coefficients: \\
+\beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
+b_j \sim N(0, \sigma_b)~~[penalized] \\
 \sigma_b \sim N(0, 0.5) (seasonal~smooth) \\
@@
 Meanwhile, this dose presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.
+This pattern implies a potential advantage and drawback of applying a high passivation dose, followed by a lower maintenance dose, of orthophosphate.
+That is, a passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms discussed elsewhere [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is effective.
 

### 2022-07-21 Rmarkdown/paper.Rmd
edits to intro and new figure 1

@@
 ---
+
+<!-- to do: fix figure order -->
 
@@
 But pipe loop study data are inherently complex and there is a real risk of misinterpreting them.
-In particular, (1) non-linearity, (2) censoring, (3) autocorrelation, (4) irregular sampling frequency, and (5) outliers are difficult to model using standard approaches (Figure 1).
+In particular, (1) non-linearity, (2) censoring, (3) autocorrelation, (4) irregular sampling frequency, and (5) outliers are difficult to model using standard approaches (Figure 1a).
 
-```{r figure-1, fig.cap="**Figure 1.** An example time series from the study described in this paper; labels 1--5 represent features of the series that represent statistical modeling challenges."}
+```{r figure-1, fig.cap="**Figure 1.
+(a)** An example time series from the study described in this paper; labels 1--5 represent features of the series that represent statistical modeling challenges.
+**(b)** Summary of the five components comprising the model fitted to pipe loop data."}
 knitr::include_graphics(here::here("Rmarkdown/figures/figure-1.png"))
@@
 The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring).
+Imputation methods such as regression on order statistics [@helsel_statistics_2012] may be useful, but in a time series the missing data may be correlated, and imputation in the context of a full data model is usually preferable [@hopke_multiple_2001].
 
@@
 It is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time, but it is not discussed in popular guides for pipe loop data analysis [@wysock_statistical_1995; @kirmeyer1996corr].
-Moreover, neither the t-test, its generalizations, nor its non-parametric analogues account for autocorrelation explicitly [@helsel_statistical; @raadt_pressing_2019].
+Moreover, neither the t-test nor its non-parametric analogues account for autocorrelation explicitly [@helsel_statistical; @raadt_pressing_2019].
 
@@
 
-The data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given their fairly rigid assumptions.
+The primary goal of a pipe loop study is to compare two or more corrosion control treatments [@wysock_statistical_1995].
+But the data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given their fairly rigid assumptions.
 Often, a more flexible approach is required, and Bayesian methods are inherently flexible, since there is no need to derive the sampling distribution of an estimator [@kruschke_doing_2015].
 
-Here, we propose a Bayesian approach for building statistical models of pipe loop study data that is sufficiently flexible for most scenarios.
-The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components:
+Here, we propose a Bayesian approach for building statistical models of pipe loop study data that is sufficiently flexible for most pipe loop datasets.
+And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions.
+The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components (Figure 1b):
 
 1. An average representing each pipe
-2. A global trend describing long-term variation in the parameter of interest across all pipes
-3. A set of pipe-specific trends that describe the deviations from the global long-term trend particular to each individual pipe
+2. A global trend describing multi-year variation in the parameter of interest across all pipes
+3. A set of pipe-specific trends that describe the deviations from the global multi-year trend particular to each individual pipe
 4.
 A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it)
-5. An error autocorrelated term
+5. An autocorrelated error term
 
-Separation of the time-trend into long-term and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the long-term trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are longer data gaps the series representing one or more pipes.
+Separation of the time-trend into long-term and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are longer data gaps the series representing one or more pipes.
 In these instances, the model "borrows" information from the global term.
@@
 Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.
-An advantage of our Bayesian approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values.
+
+We use the model to (1) compare corrosion control treatments in terms of dissolved and particulate lead, (2) compare their relative contributions to total lead release under each treatmenet, and (3) identify periods of stable lead release by considering the rate of change in the global multi-year term.
+
+An advantage of our approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values.
 This allows for full propagation of uncertainty through all stages of post-processing and analysis [@beck_multi-scale_2022].
@@
 Lead, iron, aluminum, and manganese were measured by ICP-MS [@sm_3125] with method detection limits estimated at 0.2, 5, 5, and 2 µg L^-1^, respectively.
-An aliquot of each sample was filtered at 0.45 µm using a teflon filter and a vacuum filter apparatus.
+An aliquot of each sample was filtered at 0.45 µm using a polytetrafluoroethylene (PTFE) filter and a vacuum filter apparatus.
+
+*Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high.
+In hard waters such as the feedwater for the model system, this bias is likely to be less than X%.*
+
 Temperature was measured using an infrared thermometer, total chlorine [@sm_4500_cl_g] and orthophosphate [@sm_4500_p_e] were measured using a spectrophotometer, and pH was measured using a benchtop meter.
@@
 
-The models included one-dimensional global smooths describing long-term variation in lead across all time series (Figure 4b).
-The model of dissolved lead also included a global smooth describing seasonal variation (Figure 4a).
-Both models included group-level smooths describing long-term variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019].
-The long-term smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
+The models included one-dimensional global smoothers describing multi-year variation in lead across all time series (Figure 4b).
+The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a).
+Both models included group-level smoothers describing multi-year variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019].
+The multi-year smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
 To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.
+To predict total lead, the particulate and dissolved predictions can be summed.
 
 Time series representing lead release to drinking water tend to feature extreme values caused by particulate or colloidal lead.
-These can severely bias parameter estimates unless they are accounted for.
-And since our dataset was impacted by extreme lead release events, we used a Student *t* likelihood in place of the usual Gaussian likelihood to limit the influence of outliers [@brms; @brms_adv].
+Even in 0.45 µm filtrate (i.e., dissolved lead), small colloids can sometimes elevate lead concentrations well above solubility limits [@king_role_2022].
+These "outliers" can severely bias parameter estimates unless they are accounted for.
+And since our dataset was impacted by extreme lead release events, we used a Student *t* likelihood in place of the usual Gaussian likelihood to limit the influence of extreme values [@brms; @brms_adv].
 
@@
 
-The transformed data, then, ($y_t$) were modeled as follows:
+The transformed dissolved or particulate lead time series, $y_t$, were modeled as follows:
 
@@
 Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$, where $x$ and $y$ are the lower and upper limits, respectively.
-Neither likelihood function requires imputation: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.
+Parameters, then, are estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.
 
-The local intercepts $\alpha_{pipe_i}$ (i.e., the pipe-specific average response) are defined as follows:
+The group-level intercepts $\alpha_{pipe_i}$ (i.e., the pipe-specific average responses) are defined as follows:
 
@@
 While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4).
-That is, extreme and sustained deviations from typical behavior are generally captured by the error term.
+That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but they are still reflected in model predictions: see Figure 6 and equation 5).
 
@@
 
-## Identifying periods of significant change
+## Estimating the particulate fraction
+
+Across all pipe sections, a majority of total lead was present in 0.45 µm filtrate (Figure 9).
+Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible.
+During the conditioning period---before orthophosphate was introduced---particle release depended mainly on the pipe configuration, with galvanically-coupled lead and copper ("Pb-Cu") yielding the greatest particulate fraction.
+Almost immediately after conditioning, the particulate fraction increased in the first two arrays as orthophosphate decreased lead solubility.
+The inverse seasonality---higher particulate fractions in winter---is a function of the relatively constant particulate lead release and the strongly seasonal dissolved lead release (Figure 6).
+The particulate fraction was also higher in the effluent from pipes featuring a galvanic junction.
+This, and the effect of orthophosphate on lead speciation, is consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
+
+```{r figure-9, fig.cap="**Figure 9.** The particulate fraction of total lead over time, grouped by pipe and orthophosphate dosing protocol, and estimated from the posterior distributions of the models fitted to particulate and dissolved lead time series.
+The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections."}
+knitr::include_graphics(here::here("Rmarkdown/figures/figure-9.png"))
+```
+
+## Identifying stability
 
@@
 The true seasonal relationship is probably complex and difficult to determine from the study data, but both temperature and influent iron concentration varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016].
-
-```{r figure-9, fig.cap="**Figure 9.** The particulate fraction of total lead over time, grouped by pipe and orthophosphate dosing protocol, and estimated from the posterior distributions of the models fitted to particulate and dissolved lead time series.
-The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections."}
-knitr::include_graphics(here::here("Rmarkdown/figures/figure-9.png"))
-```
-
-## Estimating the particulate fraction
-
-Across all pipe sections, a majority of total lead was present in 0.45 µm filtrate (Figure 9).
-Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible.
-During the conditioning period---before orthophosphate was introduced---the particulate fraction depended mainly on the pipe configuration.
-Almost immediately after conditioning, the particulate fraction increased in the first two arrays as orthophosphate decreased lead solubility.
-The inverse seasonality---higher particulate fractions in winter---is a function of the relatively constant particulate lead release and the strongly seasonal dissolved lead release (Figure 6).
-The particulate fraction was also higher in the effluent from pipes featuring a galvanic junction.
-This, and the effect of orthophosphate on lead speciation, is consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
 

### 2022-07-21 R/figures.R

@@ -26,7 +26,7 @@ library("lubridate", include.only = c("year", "yday"))
 library("ggtext", include.only = c("element_markdown", "geom_richtext"))
 library("tibble", include.only = c("tribble", "rowid_to_column"))
 
-theme_set(theme_bw() + theme(legend.position = "bottom"))
+theme_set(theme_bw() + theme(legend.position = "bottom", plot.tag = element_text(face = "bold")))
 palette <- wesanderson::wes_palette("Zissou1", 6, "continuous")
 
 #------------------ load ------------------
@@ -38,29 +38,50 @@ pk_vals <- readr::read_csv("data-clean/pareto-k-values.csv")
 locations <- model_in %>%
   distinct(location, ortho_dose, pipe_material)
 
+#------------------ model predictions ------------------
+
+preds_diss <- add_pred_draws_car1(model_in, model_diss)
+preds_part <- add_pred_draws_car1(model_in, model_part)
+
+preds_diss_sum <- preds_diss %>%
+  select(.epred) %>%
+  summarize_preds(y_var = model_in$lead_dissolved, recensor = TRUE)
+
+preds_part_sum <- preds_part %>%
+  select(.epred) %>%
+  summarize_preds(y_var = model_in$lead_part, recensor = TRUE)
+
 #------------------ figure 1 ------------------
+
+# panel (a):
 
 annotations <- tibble::tribble(
          ~date, ~value,                                                        ~label,
   "2018-12-01",   0.01,                                            "1. Non-linearity",
   "2020-10-01",  1e-04,                                           "2. Left-censoring",
-  "2020-03-01",   0.04, "3. Autocorrelation\n(clustering of similar\nvalues in time)",
-  "2018-04-01",    0.7,                            "4. Irregular sampling\nfrequency",
-  "2021-05-01",    0.5,                                          "5. Extreme\nvalues"
+  "2020-03-01",  0.045,     "3. Autocorrelation\n(similar values\nclustered in time)",
+  "2018-05-01",   0.85,                            "4. Irregular sampling\nfrequency",
[…]

### 2022-07-22 Rmarkdown/paper.Rmd
added fig s9, response to reviewer 3

@@
 library("stringr")
+library("janitor")
 model_in <- readr::read_csv(here::here("data-clean/model_in.csv"))
@@
 4.
-A global seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it)
+A seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it)
 5. An autocorrelated error term
 
-Separation of the time-trend into long-term and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are longer data gaps the series representing one or more pipes.
+Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are longer data gaps the series representing one or more pipes.
 In these instances, the model "borrows" information from the global term.
@@
 An aliquot of each sample was filtered at 0.45 µm using a polytetrafluoroethylene (PTFE) filter and a vacuum filter apparatus.
+Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high.
+In hard waters such as the feedwater for the model system, this bias is likely to be much lower than in soft waters.
 
-*Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high.
-In hard waters such as the feedwater for the model system, this bias is likely to be less than X%.*
+## Summary
+
+```{r data-summary}
+
+pl_summ <- model_in %>% 
+  filter(pipe_material != "Cu/Pb-Sn") %>% 
+  mutate(
+    pipe = str_extract(pipe_material, "[^\\s]+"),
+    value = 1e3 * lead # convert to µg/L
+  ) %>% 
+  group_by(pipe) %>% 
+  summarize(
+    med = median(value),
+    lq = quantile(value, .1),
+    uq = quantile(value, .9)
+  ) %>% 
+  ungroup() %>% 
+  mutate(across(where(is.double), ~ signif(.x, 2))) %>% 
+  pivot_wider(names_from = pipe, values_from = where(is.numeric)) %>% 
+  janitor::clean_names()
+
+ts_len <- model_in %>% 
+  group_by(location) %>% 
+  summarize(n = n()) %>% 
+  ungroup() %>% 
+  summarize(min = min(n), max = max(n))
+
+```
+
+Median lead in full and partial lead pipe sections (i.e., "Pb #1--2" and "Pb-Cu") was `r pl_summ$med_pb` (`r pl_summ$lq_pb`--`r pl_summ$uq_pb`) and `r pl_summ$med_pb_cu` (`r pl_summ$lq_pb_cu`--`r pl_summ$uq_pb_cu`) µg L^-1^, respectively (10^th^--90^th^ percentiles in parentheses).
+Time series representing effluent from individual pipes ranged in length from `r ts_len$min` to `r ts_len$max` date-value pairs spanning `r with(model_in, as.numeric(difftime(max(date), min(date))))` days.
 
 Temperature was measured using an infrared thermometer, total chlorine [@sm_4500_cl_g] and orthophosphate [@sm_4500_p_e] were measured using a spectrophotometer, and pH was measured using a benchtop meter.
-Water quality---including alkalinity [@sm_2320b], true colour [@sm_2120c], conductivity [@sm_2510b], fluoride [@sm_4500], total hardness [@sm_2340], and turbidity [@sm_2130b] ---is summarized in Table S4.
+Water quality---including alkalinity [@sm_2320b], true colour [@sm_2120c], conductivity [@sm_2510b], fluoride [@sm_4500], total hardness [@sm_2340], and turbidity [@sm_2130b] ---is summarized in Table S4 (all Standard Methods).
 Colour was quantified after filtration at 0.45 µm, using mixed cellulose ester membrane filters.
@@
 $$(1)~~y = \beta_{0} + f(x_t) + \epsilon_t$$ 
-where $f(x_t)$ is a smooth function (i.e., a *smooth*) of $x_t$,
+where $f(x_t)$ is a smooth function of $x_t$,
 
@@
 Particulate lead concentrations less than the detection limit of 0.2 µg L^-1^ were left-censored, and particulate lead concentrations representing the difference between an observed (total) and left-censored (dissolved) concentration were interval-censored.
-The lower bound of the interval was set as the difference between the total lead concentration and the detection limit and the upper bound as the total lead concentration.
+The upper bound of the interval was set as the total lead concentration, implying zero dissolved lead.
+The lower bound was set as the total lead concentration less the detection limit, implying a dissolved lead concentration equal to the detection limit.
 Given their impractically high censoring rates (up to 96%), time series representing effluent from soldered copper pipe sections were not modeled.
@@
 (a)** Seasonal component of the trend in dissolved lead.
-**(b)** Global long-term component of the trend in dissolved and particulate lead.
-**(c)** Group-level (i.e., pipe-specific) long-term trends in dissolved and particulate lead, representing the deviations of each time series from the global long-term trend.
+**(b)** Global multi-year component of the trend in dissolved and particulate lead.
+**(c)** Group-level (i.e., pipe-specific) multi-year trends in dissolved and particulate lead, representing the deviations of each time series from the global multi-year trend.
 In panels **(a--b)**, shaded regions span the middle 95% of the posterior distribution, and 200 draws from that posterior are superimposed in blue.
@@
 Both models included group-level smoothers describing multi-year variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019].
-The multi-year smooths were thin-plate regression splines and the seasonal smooth was a cyclic cubic regression spline [@mgcv].
+The multi-year smoothers were thin-plate regression splines and the seasonal smoother was a cyclic cubic regression spline [@mgcv].
 To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.
@@
 The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$.
-The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smooths takes the following form:
+The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smoothers takes the following form:
 
@@
 b_j \sim N(0, \sigma_b)~~[penalized] \\
-\sigma_b \sim N(0, 0.5) (seasonal~smooth) \\
-\sigma_b \sim T(0, 2.5, 3) (all~other~smooths) \\ 
+\sigma_b \sim N(0, 0.5) (seasonal~smoother) \\
+\sigma_b \sim T(0, 2.5, 3) (global,~group-level~smoothers) \\ 
 \end{align}
@@
 
-where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global long-term trend).
-The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooths [@mgcv].
+where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global multi-year trend).
+The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooth terms [@mgcv].
 And while the chosen priors did help with model convergence, they did not exert a strong influence on the posterior (Figure S2); the likelihood---representing `r nrow(model_in)` observations in each model---was overwhelmingly influential.
@@
 
-We estimated the local slope of the global long-term smooths using a finite difference method.
+We estimated the first derivative of the global multi-year smoothers using a finite difference method.
 Specifically, we generated 4000 posterior predictions of the smooth trend along a regular sequence spanning the date range of the data.
@@
 
-Time series of particulate and dissolved lead featured prominent negative trends in the first year of operation followed by an extended period of relative stability.
+Time series of particulate and dissolved lead featured prominent negative trends after the introduction of orthophosphate, followed by an extended period of relative stability.
 The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a).
-To estimate the time-varying mean representing each series, we modeled lead as the sum of a global long-term smooth, a global seasonal smooth (dissolved lead only), a group-level long-term smooth, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).
+To estimate the time-varying mean representing each series, we modeled lead as the sum of a global multi-year smoother, a seasonal smoother (dissolved lead only), a group-level multi-year smoother, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).
 
@@
 
-Accounting for autocorrelation also improved the predictive accuracy of the models.
-Approximate leave-one-out cross-validation suggests that models with continuous-time autoregressive errors yielded better predictions than otherwise identically-specified models with independent errors.
+Accounting for autocorrelation also improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation.
 Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016].
-The difference between the expected log pointwise predictive densities of the dissolved and particulate lead models, respectively, was 206 and 47.
-In both comparisons, the autoregressive error model was more accurate.
+We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors.
+The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).
 
@@
 Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
-This general pattern is represented in our model by the global long-term trend (Figure 4b), which excludes variation in effluent lead due to non-constant inputs (e.g., water quality and temperature).
+This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation.
 The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term.
 Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
-Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to changes in water quality over the study.
+Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to nonseasonal changes in water quality over the study.
 
 ```{r figure-8, fig.cap="**Figure 8.
-(a) ** The global long-term smooth and its estimated local slope.
-A random sample of 200 smooths/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions.
+(a) ** The global multi-year smoother (*f*~global~(t)) and its estimated first derivative (*f'*~global~(t)).
+A random sample of 200 curves/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions.
 The median of the global trend is shaded red when the 95% credible interval around the local slope does not include zero.
@@
 
-The variation in dissolved lead captured in Figure 8a neglects seasonality.
+The variation in dissolved lead captured in Figure 8a neglects seasonality by design.
 And while the seasonal component of the model (Figure 8b) is described as a function of ordinal day, this is a proxy for the physical and chemical inputs that are expected to have caused it.
-The true seasonal relationship is probably complex and difficult to determine from the study data, but both temperature and influent iron concentration varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016].
+The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016].
+And while dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).
 

### 2022-07-22 R/figures.R

@@ -12,6 +12,7 @@ library("purrr")
 library("patchwork")
 library("readr")
 library("posterior")
+library("lubridate")
 library("withr", include.only = "with_seed")
 library("ggdist", include.only = c("stat_halfeye", "median_qi"))
 library("glue", include.only = "glue")
@@ -22,7 +23,6 @@ library("wesanderson", include.only = "wes_palette")
 library("ggh4x", include.only = "facetted_pos_scales")
 library("mgcv", include.only = "gam")
 library("broom", include.only = "tidy")
-library("lubridate", include.only = c("year", "yday"))
 library("ggtext", include.only = c("element_markdown", "geom_richtext"))
 library("tibble", include.only = c("tribble", "rowid_to_column"))
 
@@ -1086,7 +1086,10 @@ fig8a <- slopes_sum %>%
     rows = vars(name = fct_relevel(name, "Smooth", after = 0L)),
     cols = vars(fraction),
     scales = "free_y",
-    labeller = labeller(name = as_labeller(c("Local slope" = "Local\nslope", "Smooth" = "Trend")))
+    labeller = labeller(name = as_labeller(c(
+      "Local slope" = "*f'*<sub>global</sub>(t)",
+      "Smooth" = "*f*<sub>global</sub>(t)"
+    )))
   ) +
   geom_ribbon(aes(ymin = .lower, ymax = .upper), alpha = .5) +
   geom_line(
@@ -1113,7 +1116,10 @@ fig8a <- slopes_sum %>%
     col = "grey"
   ) +
   labs(x = NULL, y = "Effect") +
-  theme(axis.text.x = element_text(angle = 35, hjust = 1))
+  theme(
+    axis.text.x = element_text(angle = 35, hjust = 1),
+    strip.text.y = ggtext::element_markdown()
+  )
 
 fig8b_in <- pdat %>%
   filter(
@@ -1129,7 +1135,7 @@ fig8b_in <- pdat %>%
   bind_rows(
     smooth_terms_diss[["mu: s(date_yday,bs=\"cc\")"]] %>%
       mutate(
-        param = "GAM (seasonal component)",
+        param = "*f*<sub>seasonal</sub>(t)",
         yday = 365 * date_yday
       )
   ) %>%
@@ -1140,10 +1146,7 @@ fig8b_in <- pdat %>%
 
 fig8b <- fig8b_in %>%
[…]

### 2022-07-25 Rmarkdown/paper.Rmd
edits to conclusion

@@
 
-We demonstrate our approach using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols.
+We demonstrate the model using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols.
 Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.
@@
 
-Time series of particulate and dissolved lead featured prominent negative trends after the introduction of orthophosphate, followed by an extended period of relative stability.
+Time series of particulate and dissolved lead featured prominent negative trends after the introduction of orthophosphate, followed by an extended period of relative stability (Figure 6).
 The dissolved lead time series were seasonal, with maximum levels in late summer and minimum levels in early spring (Figure 4a).
-To estimate the time-varying mean representing each series, we modeled lead as the sum of a global multi-year smoother, a seasonal smoother (dissolved lead only), a group-level multi-year smoother, a local intercept for each pipe, and a continuous-time autocorrelated error term (Figure 6).
-
-Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$.
-The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov.
-2020 ("Pb-Cu", Figure 6).
-While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4).
-That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but they are still reflected in model predictions: see Figure 6 and equation 5).
-
-Accounting for autocorrelation also improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation.
-Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016].
-We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors.
-The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).
 
@@
 
-Posterior predictions from the models---$\mu_t$ in Equation $(5)$---can be compared to contrast the orthophosphate dosing protocols.
+Posterior predictions from the models---$\mu_t$ in Equation $(5)$---can be compared to evaluate the orthophosphate dosing protocols.
 Here we compared the 0--0.5 mg P L^-1^ with the 1 mg P L^-1^ and the 1 mg P L^-1^ with the 2--0.75 mg P L^-1^ dosing protocols over time by considering their ratios.
 
-The three protocols yielded distinct lead release distributions over most of the study, and the difference between the lowest orthophosphate dose (0.5 mg P L^-1^) and the other two is particularly apparent (Figure 7).
-The ratio of dissolved lead concentrations in the absence of orthophosphate to those representing the 1 mg P L^-1^ dose peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a); the ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^, but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 for the majority of the study.
-This pattern is consistent with expectations based on geochemical solubility models [@schock1996corr].
+The three dosing protocols yielded distinct lead release distributions over most of the study, and the difference between the lowest orthophosphate dose and the other two is particularly apparent (Figure 7).
+The ratio of dissolved Pb concentrations at a P ratio of 0:1---comparing the 0 mg P L^-1^ control with the 1 mg P L^-1^ dose---peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a).
+The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception.
+This highlights the superior corrosion control performance of 1 mg P L^-1^ over both 0 and 0.5 mg P L^-1^ and is consistent with expectations based on geochemical solubility models [@schock1996corr].
 
-The ratio comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b).
-As a comparison of either particulate or dissolved lead concentrations, it was less than one over much of the first part of the series and greater than one over much of the latter part.
-The transition occurred a short time after orthophosphate was decreased from 2--0.75 mg P L^-1^, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose, likely as particles and small colloids [@trueman_characterizing_2019; @li_impact_2020].
-Meanwhile, this dose presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.
-This pattern implies a potential advantage and drawback of applying a high passivation dose, followed by a lower maintenance dose, of orthophosphate.
-That is, a passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms discussed elsewhere [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is effective.
+Pb ratios comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b).
+As a comparison of either particulate or dissolved lead concentrations, the Pb ratio at a 1:2 P ratio was less than one, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose.
+This likely occurred in the dissolved fraction as small colloids [@trueman_characterizing_2019; @li_impact_2020; @lytle_lead_2020].
+
+The transition from 2--0.75 mg P L^-1^ reversed this pattern: the dissolved and particulate Pb ratios at the 1:0.75 P ratio were largely greater than one.
+A high dose of 2 mg P L^-1^, then, presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.
+This implies a potential advantage and drawback of a high passivation dose, followed by a lower maintenance dose, of orthophosphate.
+That is, a passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is either more effective or effective sooner than it otherwise might be.
 
@@
 ```
+
+## Modeling autocorrelation and assessing performance
+
+Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$.
+The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov.
+2020 ("Pb-Cu", Figure 6).
+While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4).
+That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but they are still reflected in model predictions: see Figure 6 and equation 5).
+
+Accounting for autocorrelation also improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the complete dataset.
+Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016].
+We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors.
+The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).
 
@@
 Moreover, the model accounts for the reduced information content in autocorrelated observations.
-We demonstrate it using data from a four-year pipe loop study and we use the model to compare orthophosphate dosing protocols and analyze time trends.
+
+Our approach can be modified to suit the needs of diverse analyses.
+While time is the sole independent variable here, water quality parameters can easily be added as predictors if inferences concerning their effects are of interest.
+Moreover, any parameter in the likelihood can be modeled---including the standard deviation if model residuals are heteroskedastic.
+The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical].
+But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time applications.
+
+We demonstrate it here using data from a four-year pipe loop study and we use the model to compare orthophosphate dosing protocols and analyze time trends.
 The additive smooth terms comprising the model yield an informative decomposition of pipe loop time series and a more nuanced comparison of corrosion control treatments.
 
-The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the ratio difference between the 0 and 1 mg P L^-1^ dose.
-The effect of orthophosphate dose was complex: while the 1 mg P L^-1^ dose resulted in significantly lower lead release than the 0.5 mg P L^-1^ dose, the intermediate dose (0.75 mg P L^-1^) often yielded lower predicted lead levels in the latter half of the study, which we attribute to an initial high dose of 2 mg P L^-1^.
-Consistent with previous work, orthophosphate tended to increase the particulate fraction of total lead.
+The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the large difference in lead release between the 0 and 1 mg P L^-1^ dose.
+The effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^.
+We attribute the superior performance of the 0.75 mg P L^-1^ dose to an initial high dose of 2 mg P L^-1^.
+On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy.
+And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose is the conservative choice.
 The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring.

### 2022-07-27 Rmarkdown/paper.Rmd
remove fig 9

@@
 
-<!-- to do: fix figure order -->
+<!-- to do: check figure order -->
+<!-- remove figure 9 from figures.R-->
 
@@
 library("dplyr")
+library("tidyr")
 library("stringr")
@@
 Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for water quality management.
-But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common features that make popular methods, such as the *t*- or rank-sum tests, poor descriptive models.
-Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model characteristics: a robust Bayesian generalized additive model with continuous-time autoregressive errors.
+But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common attributes that make popular methods, such as the *t*- or rank-sum tests, poor descriptive models.
+Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model attributes: a robust Bayesian generalized additive model with continuous-time autoregressive errors.
 Our approach facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers.
@@
 
-```{r figure-1, fig.cap="**Figure 1.
-(a)** An example time series from the study described in this paper; labels 1--5 represent features of the series that represent statistical modeling challenges.
-**(b)** Summary of the five components comprising the model fitted to pipe loop data."}
-knitr::include_graphics(here::here("Rmarkdown/figures/figure-1.png"))
-```
-
 ## Non-linearity
@@
 
-Censoring typically arises in pipe loop studies when analyte concentrations fall below below the detection limit (i.e., left-censoring).
+Censoring typically arises in pipe loop studies when analyte concentrations fall below the detection limit (i.e., left-censoring).
 Left-censoring is often addressed by substituting a constant for values below the detection limit and sometimes by removing the censored values.
@@
 The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring).
-Imputation methods such as regression on order statistics [@helsel_statistics_2012] may be useful, but in a time series the missing data may be correlated, and imputation in the context of a full data model is usually preferable [@hopke_multiple_2001].
+Imputation methods such as regression on order statistics may be useful [@helsel_statistics_2012], but in a time series the censored data may be correlated, and imputation in the context of a full data model is usually preferable [@hopke_multiple_2001].
 
@@
 For example, the effect of a perturbation to corrosion scale is evident for some time after the fact.
-From a statistical perspective, autocorrelation makes parameter estimates more uncertain [@helsel_statistical] by reducing the effective sample size (i.e., the information content of the sample) [@gelman_bayesian_2014].
-It is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time, but it is not discussed in popular guides for pipe loop data analysis [@wysock_statistical_1995; @kirmeyer1996corr].
+From a statistical perspective, autocorrelation makes parameter estimates more uncertain by reducing the effective sample size (i.e., the information content of the sample) [@helsel_statistical; @gelman_bayesian_2014].
+As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel_statistical].
+Autocorrelation is very likely to be a feature of any pipe loop study dataset where effluent from pipes is sampled repeatedly over time, but it is not discussed in popular guides for pipe loop data analysis [@wysock_statistical_1995; @kirmeyer1996corr].
 Moreover, neither the t-test nor its non-parametric analogues account for autocorrelation explicitly [@helsel_statistical; @raadt_pressing_2019].
 
-Autocorrelation overstates the precision of statistical estimates [@gelman_bayesian_2014].
-As a consequence, it inflates the probability of falsely rejecting null hypotheses [@helsel_statistical].
 The effect of autocorrelation can be surprisingly large: 100 observations from a time series may provide no greater precision than 10 independent observations if the 100 are strongly correlated in time [@helsel_statistical].
@@
 Extreme lead or copper concentrations---typically due to particle release---are often inconsistent with the normality assumption common to many popular statistical tests.
-Outliers are sometimes omitted [@schock_causes_1990], but unless they represent measurement error or a separate population (e.g., a process error), it is usually preferable to use a statistical method that is robust against them [@helsel_statistical].
+Outliers are sometimes omitted in practice [@schock_causes_1990], but unless they represent measurement error or a separate population (e.g., a process error), it is usually preferable to use a statistical method that is robust against them [@helsel_statistical].
+
+```{r figure-1, fig.cap="**Figure 1.
+(a)** An example time series from the study described in this paper; labels 1--5 denote attributes of the series presenting statistical modeling challenges.
+**(b)** Summary of the five components comprising the model fitted to pipe loop data."}
+knitr::include_graphics(here::here("Rmarkdown/figures/figure-1.png"))
+```
 
@@
 
-Here, we propose a Bayesian approach for building statistical models of pipe loop study data that is sufficiently flexible for most pipe loop datasets.
-And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions.
+Here, we propose a Bayesian generalized additive model for pipe loop study data that is sufficiently flexible for most pipe loop datasets.
 The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components (Figure 1b):
 
-1. An average representing each pipe
-2. A global trend describing multi-year variation in the parameter of interest across all pipes
-3. A set of pipe-specific trends that describe the deviations from the global multi-year trend particular to each individual pipe
-4.
-A seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it)
-5. An autocorrelated error term
+1. An average representing each pipe.
+2.
+A smooth seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (this component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it).
+3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes.
+4. A set of smooth pipe-specific trends that describe the deviations from the global multi-year trend particular to each pipe.
+5. An autocorrelated error term.
 
-Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are longer data gaps the series representing one or more pipes.
+Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are long data gaps in the series representing one or more pipes.
 In these instances, the model "borrows" information from the global term.
-The autocorrelated error model improves the accuracy of parameter distributions and helps to prevent overfitting.
+The autocorrelated error model improves the accuracy of parameter distributions and predictions, while helping to prevent overfitting [@simpson_modelling_2018].
+And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions.
 
 We demonstrate the model using the data from a four-year pipe loop study that included multiple pipe configurations and orthophosphate dosing protocols.
-Our model estimates autocorrelation, despite an irregular sampling frequency; it also accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.
+Our model accommodates censored observations without imputation and is robust against extreme analyte concentrations due to particle release---meaning that no special treatment of outliers is necessary.
 
-We use the model to (1) compare corrosion control treatments in terms of dissolved and particulate lead, (2) compare their relative contributions to total lead release under each treatmenet, and (3) identify periods of stable lead release by considering the rate of change in the global multi-year term.
+We use the model to (1) compare corrosion control treatments in terms of dissolved and particulate lead, (2) compare their relative contributions to total lead release, and (3) identify periods of stable lead release by considering the rate of change in the global multi-year term.
 
@@
 
-## Summary
-
 ```{r data-summary}
@@
 
-Comparisons of corrosion control treatments are often aided by a statistical model.
-Typical choices in pipe loop studies include constant (*e.g.*, *t*-test) and linear models.
-But when applied to an example time series from our dataset (Figure 2a), both of these models violate a key assumption: that residuals are independent.
+Comparisons of corrosion control treatments are often aided by a statistical model, and typical choices in pipe loop studies include constant (*e.g.*, *t*-test) and linear models.
+But when applied to an example time series from our dataset (Figure 2a), both of these models violate a key assumption: the independence of model residuals.
 
@@
 This can be extended to include the residual at time *t-k*, $\epsilon_{t-k}$, where *k = 1...21*, yielding the autocorrelation function (Figure 2c).
-Independent residuals yield autocorrelation functions that hover around zero, whereas correlated residuals typically yield autocorrelation functions that either decay or oscillate.
+Independent residuals yield autocorrelation functions that remain close to zero, whereas correlated residuals yield autocorrelation functions exhibiting a pattern: decay or oscillation, for instance.
 We assume here that the example time series is a regular sequence, but in this case, that is only an approximation, and we relax it in the next section.
@@
 To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.
-To predict total lead, the particulate and dissolved predictions can be summed.
+While omitted here, total lead could be predicted using the same model structure.
 
@@
 Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$, where $x$ and $y$ are the lower and upper limits, respectively.
-Parameters, then, are estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.
+Parameters, then, were estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.
 
@@
 The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$.
-The $f_j(t)$ are smooth spline functions that comprise the GAM, and each of the $n$ smoothers takes the following form:
+The $f_j(t)$ are smooth spline functions of time that comprise the GAM, and each of the $n$ smoothers takes the following form:
 
@@
 
-The posterior predictive distribution was also robust against extreme values (Figure 5d): the Pareto-*k* values shown here represent the influence of individual observations on the posterior distribution, and values below 0.7 are considered moderately influential, yielding reliable estimates of out-of-sample predictive accuracy (see *Comparing corrosion control treatments* below) [@loo; @loo_paper].
+The posterior predictive distribution was also robust against extreme values (Figure 5d): the Pareto-*k* values shown here represent the influence of individual observations on the posterior distribution, and values below 0.7 are considered moderately influential (at most), yielding reliable estimates of out-of-sample predictive accuracy (see *Comparing corrosion control treatments* below) [@loo; @loo_paper].
 Here, all Pareto-*k* values were below 0.5.
@@
 Here we compared the 0--0.5 mg P L^-1^ with the 1 mg P L^-1^ and the 1 mg P L^-1^ with the 2--0.75 mg P L^-1^ dosing protocols over time by considering their ratios.
+This highlights an advantage of the Bayesian approach---straightforward error propagation for quantities derived from the posterior distribution of the model parameters [@korner_nievergelt_bayesian_2015].
 
@@
 The ratio of dissolved Pb concentrations at a P ratio of 0:1---comparing the 0 mg P L^-1^ control with the 1 mg P L^-1^ dose---peaked at `r signif(fig7_summ$.epred, 3)` with a 95% credible interval of `r signif(fig7_summ$.lower, 3)`--`r signif(fig7_summ$.upper, 3)` (Figure 7a).
+That is, lead release at 0 mg P L^-1^ was as much as 31 times higher than lead release at 1 mg P L^-1^.
+
 The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception.
@@
 Pb ratios comparing the 1 and 2--0.75 mg P L^-1^ protocols exhibited a more nuanced pattern (Figure 7b).
-As a comparison of either particulate or dissolved lead concentrations, the Pb ratio at a 1:2 P ratio was less than one, suggesting that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose.
+As a comparison of either particulate or dissolved lead concentrations, the Pb ratio at a 1:2 P ratio was less than one, meaning that 2 mg P L^-1^ caused excess lead release relative to the 1 mg P L^-1^ dose.
 This likely occurred in the dissolved fraction as small colloids [@trueman_characterizing_2019; @li_impact_2020; @lytle_lead_2020].
 
-The transition from 2--0.75 mg P L^-1^ reversed this pattern: the dissolved and particulate Pb ratios at the 1:0.75 P ratio were largely greater than one.
-A high dose of 2 mg P L^-1^, then, presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^, despite being exposed to a lower concentration of 0.75 mg P L^-1^.
-This implies a potential advantage and drawback of a high passivation dose, followed by a lower maintenance dose, of orthophosphate.
-That is, a passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is either more effective or effective sooner than it otherwise might be.
+The transition from 2--0.75 mg P L^-1^ reversed this pattern: the dissolved and particulate Pb ratios at the 1:0.75 P ratio were largely greater than one, meaning greater lead release at the 1 mg P L^-1^ dose.
+A high dose of 2 mg P L^-1^, then, presumably caused a scale layer to form that ultimately released less lead---in both fractions---than its counterpart formed at 1 mg P L^-1^.
+This is despite being exposed to a lower orthophosphate concentration in the latter half of the study (0.75 mg P L^-1^), and it highlights a potential advantage and drawback of the passivation-maintenance dosing strategy.
+That is, a high passivation dose may cause excess lead release in the short term---likely due to particle-generating mechanisms [@lytle_lead_2020; @zhao_formation_2018]---while ensuring that a lower maintenance dose is either more effective or effective sooner than it otherwise might be.
 
@@
 While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4).
-That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but they are still reflected in model predictions: see Figure 6 and equation 5).
+That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but still reflected in model predictions: see Figure 6 and equation 5).
 
-Accounting for autocorrelation also improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the complete dataset.
+Accounting for autocorrelation improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the full dataset.
 Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016].
@@
 
-## Estimating the particulate fraction
+## Trends in particulate lead release
 
-Across all pipe sections, a majority of total lead was present in 0.45 µm filtrate (Figure 9).
-Orthophosphate treatment was associated with a larger particulate fraction, as is particularly evident in the first year of the study, where a comparison of 0, 1, and 2 mg P L^-1^ is possible.
-During the conditioning period---before orthophosphate was introduced---particle release depended mainly on the pipe configuration, with galvanically-coupled lead and copper ("Pb-Cu") yielding the greatest particulate fraction.
-Almost immediately after conditioning, the particulate fraction increased in the first two arrays as orthophosphate decreased lead solubility.
-The inverse seasonality---higher particulate fractions in winter---is a function of the relatively constant particulate lead release and the strongly seasonal dissolved lead release (Figure 6).
-The particulate fraction was also higher in the effluent from pipes featuring a galvanic junction.
-This, and the effect of orthophosphate on lead speciation, is consistent with previous work [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
-
-```{r figure-9, fig.cap="**Figure 9.** The particulate fraction of total lead over time, grouped by pipe and orthophosphate dosing protocol, and estimated from the posterior distributions of the models fitted to particulate and dissolved lead time series.
-The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections."}
-knitr::include_graphics(here::here("Rmarkdown/figures/figure-9.png"))
-```
+Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6).
+During the conditioning period in particular---before orthophosphate was introduced---particles represented a minor contribution.
+Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
+Given the prominent seasonality in dissolved lead release, the particulate fraction exhibited inverse seasonality, with higher particulate fractions in winter (Figure 6).
+The increase in particulate lead in the final weeks of the study period is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].
 
@@
 The variation in dissolved lead captured in Figure 8a neglects seasonality by design.
-And while the seasonal component of the model (Figure 8b) is described as a function of ordinal day, this is a proxy for the physical and chemical inputs that are expected to have caused it.
+And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it.
 The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016].
@@
 On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy.
-And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose is the conservative choice.
+And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose may be a better option in this case.
 The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring.

### 2022-07-27 R/figures.R

@@ -1250,83 +1250,77 @@ fig8 <- patchwork::wrap_plots(
 
 ggsave("Rmarkdown/figures/figure-8.png", fig8, width = 3.33, height = 6, dpi = 600)
 
-#------------------ figure 9 ------------------
-
-# generate ratios:
-
-preds_fdiss <- preds_diss %>%
-  ungroup() %>%
-  mutate(
-    .epred_retrans = retrans(.epred, model_in$lead_dissolved),
-    lead_part_retrans = retrans(preds_part$.epred, model_in$lead_part),
-    lead_total = .epred_retrans + lead_part_retrans,
-    fdiss = lead_part_retrans / lead_total
-  )
-
-preds_fdiss_summ <- preds_fdiss %>%
-  group_by(across(group_vars(preds_diss))) %>%
-  summarize_preds(retrans = FALSE, pred_var = "fdiss")
-
-# annotations:
-
-annotations_short <- tibble::tribble(
-            ~x, ~pipe_material,                                      ~labels,  ~y, ~ortho_dose,
-  "2018-01-01",        "Pb #1", "P introduced:<br>0-0.5 mg P L<sup>-1</sup>", Inf,     "0-0.5",
-  "2019-03-01",        "Pb #2", "P decreased:<br>2-0.75 mg P L<sup>-1</sup>", Inf,  "2.0-0.75",
-  "2017-07-01",        "Pb-Cu",     "Conditioning:<br>0 mg P L<sup>-1</sup>", Inf,          NA
-  )
-
-lines_short <- lines %>%
-  distinct(ortho_dose, x) %>%
-  filter(x < "2020-01-01", x > "2019-01-01") %>%
-  mutate(ortho_dose = str_remove(ortho_dose, " mg P L<sup>-1</sup>"))
-
-# plot:
-
-fig9 <- preds_fdiss_summ %>%
-  ggplot(aes(date, fdiss, col = ortho_dose, fill = ortho_dose)) +
-  facet_wrap(vars(pipe_material), ncol = 1) +
-  geom_rect(
-    data = function(x) {
-      x %>%
-        group_by(pipe_material, ortho_dose) %>%
-        summarize(xmin = min(date) - 50) %>%
-        mutate(
-          xmax = as.Date("2018-03-13"),
-          ymin = 0, ymax = Inf
-        )
-    },
-    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
[…]

### 2022-07-28 Rmarkdown/paper.Rmd
added responses to reviewers

@@
 <!-- to do: check figure order -->
-<!-- remove figure 9 from figures.R-->
 
@@
 
-1. An average representing each pipe.
+1. An average representing each pipe (all other components are centered on zero).
 2.
@@
 Separation of the time-trend into multi-year and seasonal components allows for easier identification of a stable phase in the presence of significant seasonality, and separation of the multi-year trend into global and pipe-specific components facilitates trend estimation in the presence of censoring or where there are long data gaps in the series representing one or more pipes.
-In these instances, the model "borrows" information from the global term.
+In these instances, the model "borrows" information from the other time series.
 The autocorrelated error model improves the accuracy of parameter distributions and predictions, while helping to prevent overfitting [@simpson_modelling_2018].
@@
 Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high.
-In hard waters such as the feedwater for the model system, this bias is likely to be much lower than in soft waters.
+In hard waters such as the feedwater for the model system, this bias is likely to be much lower than in soft waters [@minning_systematic_2015].
 
@@
 $b_j(x_t)$ are the functions of $x_t$ on which $y$ is regressed, $\beta_j$ are the model coefficients, and $\epsilon_t$ is the error term.
-We fitted the example GAM in Figures 2--3 using the R package *mgcv*, which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.
+We fitted the example GAM in Figures 2--3 using the R package *mgcv* [@mgcv], which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.
 
@@
 $$(3)~~\epsilon_t = \phi\epsilon_{t-1} + \omega_t$$
-where $\epsilon_t$ is the residual at time $t$, $\phi$ is the autoregressive parameter, and $\omega_t$ is an independent draw from a probability distribution.
+where $\epsilon_t$ is the residual at time $t$, $\phi$ is the autoregressive parameter, and $\omega_t$ is an independent draw from a continuous---typically Gaussian---probability distribution.
 
@@
 Unmodeled positive autocorrelation, as in Figure 2b-c, overstates the precision of parameter estimates, and so it should be either eliminated via model selection or taken into account explicitly.
-We chose to analyze the full length of each time series---including the conditioning period---to best estimate the autocorrelation structure.
+Here, we chose to analyze the full length of each time series---including the conditioning period---to best estimate the autocorrelation structure.
 
@@
 The model of dissolved lead also included a global smoother describing seasonal variation (Figure 4a).
+And while the model assumes that seasonality strength does not vary over time, some previous work has suggested that orthophosphate may damp seasonal variation in lead [@colling_measurement_1987].
+This effect could be incorporated by allowing the seasonal smoother to vary with orthophosphate dose.
+
 Both models included group-level smoothers describing multi-year variation in lead release from each pipe (Figure 4c), a hierarchical structure that is discussed elsewhere [@pedersen_hierarchical_2019].
@@
 To speed up computation, the response variable was log-transformed and then scaled to have zero mean and unit variance.
-While omitted here, total lead could be predicted using the same model structure.
+And while it is omitted here, total lead could be predicted using the same model structure as dissolved lead.
 
@@
 where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$.
+($Gamma$ and $N$ represent the gamma and Gaussian distributions.)
 
@@
 f_j(t) = X_j\beta_j + Z_jb_j \\
+\\
 priors~for~GAM~coefficients: \\
@@
 
-where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function associated with the global multi-year trend).
+where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function included in the global multi-year trend).
 The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooth terms [@mgcv].
@@
 
-## Modeling autocorrelation and assessing performance
+## Modeling autocorrelation
 
@@
 Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
-Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to nonseasonal changes in water quality over the study.
+Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to nonseasonal changes in water quality over the study (see *Trends in particulate lead release*).
 
@@
 The pipe loop system responded strongly to orthophosphate treatment, as evident in the decreasing global trend during the first year of operation and the large difference in lead release between the 0 and 1 mg P L^-1^ dose.
-The effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^.
+But the effect of orthophosphate dose was complex: lead release at 1 mg P L^-1^ was lower than at 0.5 mg P L^-1^ but higher than at 0.75 mg P L^-1^.
 We attribute the superior performance of the 0.75 mg P L^-1^ dose to an initial high dose of 2 mg P L^-1^.

### 2022-07-28 R/figures.R

@@ -991,14 +991,14 @@ p1 <- preds_d_summ %>%
     pipe_material %in% c("Pb #1", "Pb #2"),
     difference == "0-0.5 mg P L<sup>-1</sup> /<br> 1.0 mg P L<sup>-1</sup>"
   ) +
-  labs(y = expression(frac("0-0.5 mg P L"^-1, "1 mg P L"^-1)))
+  labs(y = expression(frac("[Pb]"["0-0.5 mg P L"^-1], "[Pb]"["1 mg P L"^-1])))
 
 p2 <- preds_d_summ %>%
   ratio_plot(
     pipe_material %in% c("Pb #1", "Pb #2"),
     difference == "1.0 mg P L<sup>-1</sup> /<br> 2.0-0.75 mg P L<sup>-1</sup>"
   ) +
-  labs(y = expression(frac("1 mg P L"^-1, "2-0.75 mg P L"^-1)))
+  labs(y = expression(frac("[Pb]"["1 mg P L"^-1], "[Pb]"["2-0.75 mg P L"^-1])))
 
 fig7 <- patchwork::wrap_plots(p1, p2, nrow = 2) +
   patchwork::plot_annotation(tag_level = "a") +
@@ -1021,7 +1021,7 @@ toc_art <- preds_d_summ %>%
     difference == "1.0 mg P L<sup>-1</sup> /<br> 2.0-0.75 mg P L<sup>-1</sup>"
   ) +
   theme(strip.text.y = element_blank()) +
-  labs(y = expression(frac("1 mg P L"^-1, "2-0.75 mg P L"^-1)))
+  labs(y = expression(frac("[Pb]"["1 mg P L"^-1], "[Pb]"["2-0.75 mg P L"^-1])))
 
 ggsave("Rmarkdown/figures/figure-toc.png", toc_art, width = 3.33, height = 1.8, dpi = 600)
 
@@ -1249,78 +1249,6 @@ fig8 <- patchwork::wrap_plots(
   patchwork::plot_annotation(tag_level = "a")
 
 ggsave("Rmarkdown/figures/figure-8.png", fig8, width = 3.33, height = 6, dpi = 600)
-
-# #------------------ figure 9 ------------------
-#
-# # generate ratios:
-#
-# preds_fdiss <- preds_diss %>%
-#   ungroup() %>%
-#   mutate(
-#     .epred_retrans = retrans(.epred, model_in$lead_dissolved),
[…]

### 2022-07-29 Rmarkdown/paper-si.Rmd
added fig s9, response to reviewer 3
remove fig 9
fig 7 annotations, response doc

@@
 
-This document contains 8 figures, 4 tables, and 12 pages.
+This document contains 9 figures, 4 tables, and 13 pages.
 
@@
 
+```{r figure-s9, fig.cap="**Figure S9. (a)** True colour in pipe loop feedwater, by month and year. **(b)** Aluminum in feedwater, by orthophosphate dose. "}
+knitr::include_graphics(here::here("Rmarkdown/figures/figure-s9.png"))
+```
+
 # References

### 2022-07-29 Rmarkdown/paper.Rmd
fig 7 annotations, response doc

@@
 They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995].
-The most common methods for describing pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].
+The most common models for pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].
 
@@
 Neither approach is advisable as both introduce artificial patterns that can distort results, sometimes seriously [@helsel_statistics_2012].
-Popular statistical tests (e.g., the paired t-test and its generalization, the one-way repeated measures ANOVA) cannot accommodate censoring [@helsel_statistics_2012].
+Popular statistical tests---for example, the paired t-test and its generalization, the one-way repeated measures ANOVA---cannot accommodate censoring [@helsel_statistics_2012].
 The non-parametric rank-sum test does allow for left-censoring at one detection limit, but it is not adequate for more complex censoring schemes (e.g., multiple detection limits or interval censoring).
@@
 Here, we propose a Bayesian generalized additive model for pipe loop study data that is sufficiently flexible for most pipe loop datasets.
-The collection of measurement series representing variation in a given parameter over time---often lead--- is represented as the sum of five components (Figure 1b):
+The collection of measurement series representing variation in a given parameter over time---often lead---is represented as the sum of five components (Figure 1b):
 
@@
 2.
-A smooth seasonal trend, whose strength depends on the magnitude of the seasonal variation in water quality and temperature (this component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it).
-3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes.
+A smooth seasonal trend (i.e., a function of ordinal day), whose strength depends on the magnitude of the seasonal variation in water quality and temperature.
+This component may not be necessary if there is no distinct seasonality, or if the series is too short to identify it.
+3. A smooth global trend describing multi-year variation in the parameter of interest across all pipes (i.e., a function of time).
 4. A set of smooth pipe-specific trends that describe the deviations from the global multi-year trend particular to each pipe.
@@
 
-An advantage of our approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters, and all quantities derived from those parameters, including the predicted values.
-This allows for full propagation of uncertainty through all stages of post-processing and analysis [@beck_multi-scale_2022].
-Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in lead concentrations over time.
+An advantage of our approach is the ability to move beyond point estimates and intervals: Bayesian models yield the full posterior distribution of all parameters and all quantities derived from those parameters, including the predicted values.
+This allows for propagation of uncertainty through all stages of post-processing and analysis [@beck_multi-scale_2022].
+Moreover, the use of additive smoothing splines permits a description of time series using multiple nonlinear components, which can yield a more nuanced understanding of differences in analyte concentrations over time.
 
@@
 Since the difference between model predictions and the censoring limit is not an accurate estimate of residual model error, we simulated residuals for model assessment by imputing censored values with posterior predictions and computing the residuals using the augmented dataset.
-Simulated residuals exhibited minimal autocorrelation, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation.
+Simulated residuals were minimally autocorrelated, while predictions that excluded the autocorrelation term yielded much stronger residual autocorrelation.
 The distribution of the autocorrelation coefficient for each model, $\phi$, is shown in Figure 5c, and simulated residuals as a function of time are shown in Figure S8.
@@
 Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
-Given the prominent seasonality in dissolved lead release, the particulate fraction exhibited inverse seasonality, with higher particulate fractions in winter (Figure 6).
-The increase in particulate lead in the final weeks of the study period is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].
+Given the prominent seasonality in dissolved lead release, the particulate fraction was inversely seasonal, with higher particulate fractions in winter (Figure 6).
+The increase in particulate lead in the final weeks of the study is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].
 
@@
 
-Effluent lead concentrations in pipe loop studies typically exhibit a pronounced negative trend after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
+Effluent lead concentrations in pipe loop studies often decrease sharply after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
 This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation.
@@
 Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
-Both fractions exhibited positive and negative trends in subsequent years that are difficult to explain, but they may be a response to nonseasonal changes in water quality over the study (see *Trends in particulate lead release*).
+The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to nonseasonal changes in water quality (see *Trends in particulate lead release*).
 
@@
 The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016].
-And while dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).
+And although dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).
 
@@
 
-Given that pipe loop studies are expected to remain important for drinking water quality management, time series analysis will be  necessary for reliable statistical analysis of the data they generate.
+Given that pipe loop studies are expected to remain important for drinking water quality management, time series methods will be  necessary for reliable statistical analysis of the data they generate.
 Methods that do not account for the effect of repeated measurements---and the reduced information content that results from autocorrelation---are likely to be less useful, and may even be misleading.

### 2022-07-29 R/figures.R

@@ -51,6 +51,9 @@ preds_part_sum <- preds_part %>%
   select(.epred) %>%
   summarize_preds(y_var = model_in$lead_part, recensor = TRUE)
 
+smooth_terms_part <- conditional_smooths(model_part)
+smooth_terms_diss <- conditional_smooths(model_diss)
+
 #------------------ figure 1 ------------------
 
 # panel (a):
@@ -69,7 +72,7 @@ fig1a <- model_in %>%
   filter(location == "LP31") %>%
   pivot_longer(c(lead_part, lead_dissolved), names_to = "type") %>%
   ggplot(aes(date, value)) +
-  geom_rug(sides = "t", length = unit(.02, "npc")) +
+  geom_rug(sides = "t", length = unit(.02, "npc"), size = .3) +
   geom_label(
     data = annotations,
     aes(label = label),
@@ -79,7 +82,6 @@ fig1a <- model_in %>%
   ) +
   geom_segment(
     data = tibble(x = "2019-02-01", xend = "2019-09-01", y = 1.1, yend = 2),
-    # data = tibble(x = "2018-12-01", xend = "2019-09-01", y = .72, yend = 2),
     aes(x = as.Date(x), xend = as.Date(xend), y = y, yend = yend),
     inherit.aes = FALSE, size = .3
   ) +
@@ -436,9 +438,6 @@ ggsave("Rmarkdown/figures/figure-3.png", fig3, width = 3.33, height = 4, dpi = 6
 #------------------ figure 4 ------------------
 
 n_smooths <- 200
-
-smooth_terms_part <- conditional_smooths(model_part)
-smooth_terms_diss <- conditional_smooths(model_diss)
 
 # repeated for supplementary figures of the models fitted to simulated data:
 
@@ -898,10 +897,10 @@ annotate_ratios <- tibble::tribble(
                                                    ~difference,        ~date, ~y,   ~label,
      "0-0.5 mg P L<sup>-1</sup> /<br> 1.0 mg P L<sup>-1</sup>", "2017-10-01",  0,    "0:0",
      "0-0.5 mg P L<sup>-1</sup> /<br> 1.0 mg P L<sup>-1</sup>", "2018-09-01",  0,    "0:1",
-     "0-0.5 mg P L<sup>-1</sup> /<br> 1.0 mg P L<sup>-1</sup>", "2019-09-01",  0,  "0.5:1",
[…]

### 2022-08-03 Rmarkdown/paper.Rmd
edits from coauthors

@@
 
-$$(1)~~y = \beta_{0} + f(x_t) + \epsilon_t$$ 
-where $f(x_t)$ is a smooth function of $x_t$,
+$$(1)~~y_t = \beta_{0} + f(t) + \epsilon_t$$ 
+where $\beta_0$ is the intercept, $f(t)$ is a smooth function of $t = time$,
 
-$$(2)~~f(x_t) = \sum_{j=1}^kb_j(x_t)\beta_j$$
-$b_j(x_t)$ are the functions of $x_t$ on which $y$ is regressed, $\beta_j$ are the model coefficients, and $\epsilon_t$ is the error term.
+$$(2)~~f(t) = \sum_{j=1}^kb_j(t)\beta_j$$
+$b_j(t)$ are the basis functions of $t$ that comprise $f(t)$, $\beta_j$ are the model coefficients (i.e., the basis function weights), and $\epsilon_t$ is the error term.
 We fitted the example GAM in Figures 2--3 using the R package *mgcv* [@mgcv], which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.
@@
 
-<!-- $$\begin{align} -->
-<!-- P(\theta|D) \propto P(D|\theta)P(\theta) \\ -->
-<!-- posterior \propto likelihood \times prior \\ -->
-<!-- posterior = P(\theta | D) \\ -->
-<!-- likelihood = P(D | \theta) \\ -->
-<!-- prior = P(\theta) -->
-<!-- \end{align}$$ -->
-
 The transformed dissolved or particulate lead time series, $y_t$, were modeled as follows:
@@
 likelihood: \\
-y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu) \\ 
-y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu) \\
-y_t | censored_t = 2 \sim G(\mu_t,\sigma,\nu) \\
+y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu)~~[observed] \\ 
+y_t | censored_t = 1 \sim F(\mu_t,\sigma,\nu)~~[left \text{-} censored] \\
+y_t | censored_t = 2 \sim g(\mu_t,\sigma,\nu)~~[interval \text{-} censored] \\
 \\
@@
 priors: \\
-\sigma \sim T(0, 2.5, 3) \\
+\sigma \sim Half \text{-} T(0, 2.5, 3) \\
 \nu \sim Gamma(2,0.5) \\
@@
 
-Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq x)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $x$, *at most*) [@stan2021; @mcelreath_statistical_2016].
-Interval-censored observations are indicated by $censored = 2$ and were modeled using $G(x, y)$, or $P(x \lt X \leq y) = F(y) - F(x)$, where $x$ and $y$ are the lower and upper limits, respectively.
-Parameters, then, were estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(x)$, only.
+Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq l)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $l$, *at most*) [@stan2021; @mcelreath_statistical_2016].
+Interval-censored observations are indicated by $censored = 2$ and were modeled using $g(l, u)$, or $P(l \lt X \leq u) = F_X(u) - F_X(l)$, where $l$ and $u$ are the lower and upper limits, respectively.
+Parameters, then, were estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(l)$, only.
 
@@
 \alpha \sim T(0.1, 2.5, 3)~~[prior~on~average~pipe] \\
-\sigma_{\alpha} \sim T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
+\sigma_{\alpha} \sim Half \text{-} T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
 \end{align}
@@
 b_j \sim N(0, \sigma_b)~~[penalized] \\
-\sigma_b \sim N(0, 0.5) (seasonal~smoother) \\
-\sigma_b \sim T(0, 2.5, 3) (global,~group-level~smoothers) \\ 
+\sigma_b \sim Half \text{-} N(0, 0.5) (seasonal~smoother) \\
+\sigma_b \sim  \text{-} T(0, 2.5, 3) (global/group \text{-}level~smoothers) \\ 
 \end{align}
@@
 
-The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5th percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception.
+The dissolved Pb ratio declined following the introduction of orthophosphate at 0.5 mg P L^-1^ (i.e., at a 0.5:1 P ratio), but its 2.5^th^ percentile---the lower bound of the 95% credible interval---remained above 1 with a single brief exception.
 This highlights the superior corrosion control performance of 1 mg P L^-1^ over both 0 and 0.5 mg P L^-1^ and is consistent with expectations based on geochemical solubility models [@schock1996corr].
@@
 
-## Modeling autocorrelation
-
-Autocorrelation models the dependence of observations on previous observations, or the response of the series at time $t+s$ to a shock at time $t$.
-The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov.
-2020 ("Pb-Cu", Figure 6).
-While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4).
-That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but still reflected in model predictions: see Figure 6 and equation 5).
-
-Accounting for autocorrelation improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the full dataset.
-Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016].
-We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors.
-The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).
-
-## Trends in particulate lead release
-
-Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6).
-During the conditioning period in particular---before orthophosphate was introduced---particles represented a minor contribution.
-Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
-Given the prominent seasonality in dissolved lead release, the particulate fraction was inversely seasonal, with higher particulate fractions in winter (Figure 6).
-The increase in particulate lead in the final weeks of the study is not easily explained, but it may be related to increasing feedwater aluminum concentrations (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].
-
 ## Identifying stability
 
-Effluent lead concentrations in pipe loop studies often decrease sharply after commissioning, followed by relative stability for the remainder of the study [@wysock_statistical_1995], assuming other inputs to the system remain constant.
-This general pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation.
+In a typical pipe loop study, effluent lead concentrations decrease sharply after commissioning.
+Assuming inputs to the system remain constant, a period of relative stability usually follows [@wysock_statistical_1995].
+This pattern is represented in our model by the global multi-year trend (Figure 4b), which excludes pipe-specific and seasonal water quality variation.
 The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term.
 Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
-The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to nonseasonal changes in water quality (see *Trends in particulate lead release*).
+The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to changes in water quality not captured by the model (see *Trends in particulate lead release*).
 
@@
 
-The variation in dissolved lead captured in Figure 8a neglects seasonality by design.
+The variation in dissolved lead evident in Figure 8a ignores seasonality by design.
 And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it.
 The true seasonal relationship is probably complex and difficult to determine from the study data, but temperature and influent iron concentration are plausible drivers: both varied seasonally and have been shown to impact lead release (Figure 8b--c) [@masters_increased_2015; @masters_seasonal_2016; @trueman_galvanic_2017; @trueman_understanding_2016].
-And although dissolved organic carbon has also been shown to impact lead release [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).
+And although dissolved organic carbon has also been shown to increase lead solubility [@korshin_influence_2005; @lin_release_2008; @king_role_2022], true colour---a proxy for dissolved organic carbon [@redden_chemical_2021]---was not strongly seasonal (Figure S9).
+
+## Trends in particulate lead release
+
+Over much of the study, the dissolved fraction (<0.45 µm) dominated total lead release (Figure 6).
+During the conditioning phase in particular---before orthophosphate was introduced---particles represented a minor contribution.
+Orthophosphate treatment caused a sharp decrease in dissolved lead (<0.45 µm) and a concomitant increase in the particulate fraction, which is consistent with previous observations and geochemical models [@dore_study_2019; @lytle_lead_2020; @zhao_formation_2018; @abokifa_modeling_2017; @ng_new_2018; @li_controlling_2020].
+The prominent seasonality in dissolved lead release implies an inversely seasonal particulate fraction (Figure 6).
+That is, particles contributed more to the total lead concentration in winter.
+
+The decrease and subsequent increase in particulate lead beginning in late 2020 are not easily explained, but they may be related to water temperature.
+Seasonal temperature variation was most pronounced in the final 1.5 years of the study (Figure 8c), coinciding approximately with the apparent seasonal pattern in particulate lead (Figure 8a).
+Increasing feedwater aluminum concentrations may also have been a factor (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].
+
+## Modeling autocorrelation
+
+Autocorrelation models the dependence of observations on previous observations, or the response of a measurement series at time $t+s$ to a shock at time $t$.
+The effect of the autocorrelated error term here is most evident in the response of the model to high particulate lead release after short sections of lead pipe were removed from the lead-copper pipe couples in Nov.
+2020 ("Pb-Cu", Figure 6).
+While the time-varying mean $\mu_t$ exhibits pronounced spikes corresponding to particle release events, the underlying GAM remains smooth (Figure 4).
+That is, extreme and sustained deviations from typical behavior are generally captured by the error term (but still reflected in model predictions: see Figure 6 and equation 5).
+
+Accounting for autocorrelation improved the predictive accuracy of the models, assessed here using approximate leave-one-out cross-validation on the full dataset.
+Approximate leave-one-out cross-validation estimates the expected log pointwise predictive density---a metric of model accuracy---for new observations not used to fit the model [@loo; @loo_paper; @mcelreath_statistical_2016].
+We applied it to the particulate and dissolved lead models with continuous-time autoregressive errors and also to identically-specified models with independent errors.
+The autoregressive error models yielded higher expected log pointwise predictive densities (i.e., greater accuracy), exceeding the independent error models in this metric by 206 and 47 (dissolved and particulate lead models, respectively).
 
@@
 The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical].
-But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time applications.
+But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time.
 

### 2022-08-04 Rmarkdown/paper.Rmd
revised submission

@@
 ---
-
-<!-- to do: check figure order -->
 
@@
 Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for water quality management.
-But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common attributes that make popular methods, such as the *t*- or rank-sum tests, poor descriptive models.
+But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common attributes that make popular methods, such as the t- or rank-sum tests, poor descriptive models.
 Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model attributes: a robust Bayesian generalized additive model with continuous-time autoregressive errors.
@@
 We demonstrate it using a four-year pipe loop study, with multiple pipe configurations and orthophosphate dosing protocols, finding that an initially high dose of orthophosphate (2 mg P L^-1^) that is subsequently lowered (0.75 mg P L^-1^) can yield lower lead release than an intermediate dose (1 mg P L^-1^) in the long term.
+Water utilities face difficult tradeoffs in applying orthophosphate for corrosion control, and better models of pipe loop data can help inform the decision-making process.
 
@@
 (a)** An example time series from the study described in this paper; labels 1--5 denote attributes of the series presenting statistical modeling challenges.
-**(b)** Summary of the five components comprising the model fitted to pipe loop data."}
+**(b)** Summary of the five components that comprise the proposed model."}
 knitr::include_graphics(here::here("Rmarkdown/figures/figure-1.png"))
@@
 But the data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given their fairly rigid assumptions.
-Often, a more flexible approach is required, and Bayesian methods are inherently flexible, since there is no need to derive the sampling distribution of an estimator [@kruschke_doing_2015].
+Often, a more flexible approach is required, and Bayesian methods are inherently flexible since there is no need to derive the sampling distribution of an estimator [@kruschke_doing_2015].
 
@@
 In these instances, the model "borrows" information from the other time series.
-The autocorrelated error model improves the accuracy of parameter distributions and predictions, while helping to prevent overfitting [@simpson_modelling_2018].
+The autocorrelated error model improves the accuracy of parameter distributions and predictions while helping to prevent overfitting [@simpson_modelling_2018].
 And while time is the only covariate in our model, additional predictors can easily be added as smooth or linear functions.
@@
 This was done to facilitate a biofilm study not discussed here, and the resulting total chlorine concentrations are summarized in Table S2.
-In November 2020, 30 cm sections of lead pipe were cut from the galvanic lead-copper pipes for supplementary analysis.
+In Nov.
+2020, 30 cm sections of lead pipe were cut from the galvanic lead-copper pipes for supplementary analysis.
 These sections were replaced with braided PVC hoses.
@@
 This can be extended to include the residual at time *t-k*, $\epsilon_{t-k}$, where *k = 1...21*, yielding the autocorrelation function (Figure 2c).
-Independent residuals yield autocorrelation functions that remain close to zero, whereas correlated residuals yield autocorrelation functions exhibiting a pattern: decay or oscillation, for instance.
-We assume here that the example time series is a regular sequence, but in this case, that is only an approximation, and we relax it in the next section.
+Independent residuals yield autocorrelation functions that remain close to zero, whereas correlated residuals yield autocorrelation functions that deviate from zero in a regular pattern: decay or oscillation, for instance.
+We assume here that the example time series is evenly-spaced, but in this case, that is only an approximation, and we relax it in the next section.
 
@@
 
-$$(2)~~f(t) = \sum_{j=1}^kb_j(t)\beta_j$$
-$b_j(t)$ are the basis functions of $t$ that comprise $f(t)$, $\beta_j$ are the model coefficients (i.e., the basis function weights), and $\epsilon_t$ is the error term.
+$$(2)~~f(t) = \sum_{i=1}^ns_i(t)\beta_i$$
+$s_i(t)$ are the basis functions of $t$ that comprise $f(t)$, $\beta_i$ are the model coefficients (i.e., the basis function weights), and $\epsilon_t$ is the error term.
 We fitted the example GAM in Figures 2--3 using the R package *mgcv* [@mgcv], which uses penalized likelihood maximization instead of ordinary least squares to choose coefficients while protecting against overfitting.
@@
 A reasonable estimate of $\phi$ for the linear model residuals in Figure 2b, then, would be *r*.
-Unmodeled positive autocorrelation, as in Figure 2b-c, overstates the precision of parameter estimates, and so it should be either eliminated via model selection or taken into account explicitly.
+Unmodeled positive autocorrelation, as in Figure 2b-c, yields parameter estimates that are overly precise, and so it should be either eliminated via model selection or taken into account explicitly.
 Here, we chose to analyze the full length of each time series---including the conditioning period---to best estimate the autocorrelation structure.
 
-While equation (3) is limited to time series that are regularly observed, it can be generalized to an irregular sequence of observations as a continuous-time first-order autoregression [@simpson_modelling_2018].
+And while equation (3) is limited to time series that are regularly observed, it can be generalized to an irregular sequence of observations as a continuous-time first-order autoregression [@simpson_modelling_2018].
 This more general model has the following autocorrelation function:
@@
 Autocorrelation, then, decays exponentially in time; if two observations are spaced far enough apart, they are approximately independent.
-Given the irregular spacing of observations in our data, we modeled autocorrelation using equation $(4)$.
+Given the irregular spacing of observations in our data, we modeled autocorrelation using the function defined in equation $(4)$.
 
@@
 \\
-model: \\
+model~for~\mu_t: \\
 \mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
@@
 
-Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F(x)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq l)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $l$, *at most*) [@stan2021; @mcelreath_statistical_2016].
+Left-censored observations are indicated by the variable $censored = 1$ and were modeled using $F_X(l)$, the cumulative Student *t*-distribution (*i.e.*, $P(X \leq l)$, the probability that, given a set of distributional parameters, a random variable $X$ is equal to the censoring limit, $l$, *at most*) [@stan2021; @mcelreath_statistical_2016].
 Interval-censored observations are indicated by $censored = 2$ and were modeled using $g(l, u)$, or $P(l \lt X \leq u) = F_X(u) - F_X(l)$, where $l$ and $u$ are the lower and upper limits, respectively.
-Parameters, then, were estimated without imputation of nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F(l)$, only.
+Parameters, then, were estimated without imputing nondetects: at each iteration of the MCMC algorithm, censored observations factor into the computed posterior via the cumulative probability distribution, $F_X(x)$, only.
 
@@
 \alpha_{pipe_i} \sim N(\alpha, \sigma_\alpha)~~[adaptive~prior] \\
-\alpha \sim T(0.1, 2.5, 3)~~[prior~on~average~pipe] \\
-\sigma_{\alpha} \sim Half \text{-} T(0, 2.5, 3)~~[prior~on~\sigma~of~pipes] \\ 
+\alpha \sim T(0.1, 2.5, 3)~~[prior~on~global~intercept] \\
+\sigma_{\alpha} \sim Half \text{-} T(0, 2.5, 3)~~[prior~on~\sigma~of~pipe~averages] \\ 
 \end{align}
@@
 \\
-priors~for~GAM~coefficients: \\
+priors~on~GAM~coefficients: \\
 \beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
@@
 \sigma_b \sim Half \text{-} N(0, 0.5) (seasonal~smoother) \\
-\sigma_b \sim  \text{-} T(0, 2.5, 3) (global/group \text{-}level~smoothers) \\ 
+\sigma_b \sim  Half \text{-} T(0, 2.5, 3) (global/group \text{-}level~smoothers) \\ 
 \end{align}
@@
 
-where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is $1\times1$, corresponding to a single unpenalized basis function included in the global multi-year trend).
+where $Z_j$ and $X_j$ are matrices representing the penalized and unpenalized basis functions evaluated at time $t$; and $\beta$ and $b_j$ represent the corresponding spline coefficient vectors ($\beta$ is scalar, corresponding to a single unpenalized basis function included in the global multi-year trend).
 The priors on $\sigma_b$ help to prevent overfitting by penalizing complexity in the fitted smooth terms [@mgcv].
 And while the chosen priors did help with model convergence, they did not exert a strong influence on the posterior (Figure S2); the likelihood---representing `r nrow(model_in)` observations in each model---was overwhelmingly influential.
-A summary of the fitted models, including convergence statistics and effective sample sizes, is provided in Figure S3 and S4.
+A summary of the fitted models, including convergence statistics and effective sample sizes, is provided in Figures S3 and S4.
 
@@
 (a) ** Histograms representing the median of 4000 residual draws from the models; the theoretical densities (red curves) are calculated from the posterior medians of the distributional parameters (i.e., $\\sigma$, $\\nu$) from each model.
-**(b)** Density of first-order autocorrelation, estimated using the Spearman rank correlation coefficient and 4000 residual draws from the model including and excluding the continuous-time first-order autocorrelation term (GAM and CAR(1), respectively).
+**(b)** Density of first-order autocorrelation, estimated using the Spearman rank correlation coefficient and 4000 residual draws from the model including and excluding the continuous-time first-order autocorrelation term (CAR(1) and GAM, respectively).
 Estimates were generated using an augmented dataset where censored observations were imputed via posterior prediction.
@@
 The expected value of the posterior predictive distribution (*i.e.*, $\\mu_t$ in Equation 5) is summarized by the median (blue and red lines) and the 2.5--97.5^th^ percentiles (red/blue shaded regions).
-The grey shaded region at the left side of each panel represents the conditioning period common to all pipe sections, the yellow shaded bands represent brief periods when orthophosphate dosing was interrupted, and ticks at the top of each panel represent values outside the plot limits.
-Other significant disturbances are labeled, as described in the methods section (see *Model distribution system*)."}
+The yellow vertical bands represent brief periods when orthophosphate dosing was interrupted, and ticks at the top of each panel represent values outside the plot limits.
+Orthophosphate doses are annotated."}
 knitr::include_graphics(here::here("Rmarkdown/figures/figure-6.png"))
@@
 
-```{r figure-7, fig.cap="**Figure 7.** Ratios of posterior predictions comparing **(a)** 0--0.5 with 1 mg P L^-1^ (numerator and denominator, respectively) and **(b)** 1 with 2--0.75 mg P L^-1^.
+```{r figure-7, fig.cap="**Figure 7.** Ratios of posterior predictions comparing lead concentrations at **(a)** 0--0.5 with 1 mg P L^-1^ (numerator and denominator, respectively) and **(b)** 1 with 2--0.75 mg P L^-1^.
 The bold black/red line represents the median of the posterior distribution and the shaded grey region spans the 95% credible interval; black and red indicate portions of the series where the interval included and excluded, respectively, a ratio of 1 (i.e., equality).
@@
 Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
-The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to changes in water quality not captured by the model (see *Trends in particulate lead release*).
+The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to changes in water quality not captured directly by the model (see *Trends in particulate lead release*).
 
@@
 A random sample of 200 curves/slope estimates is superimposed, and both sets of lines are underlain by shaded regions covering the middle 95% of the respective posterior distributions.
-The median of the global trend is shaded red when the 95% credible interval around the local slope does not include zero.
+The median of the global trend is shaded red when the 95% credible interval around *f'*~global~(t) does not include zero.
 **(b--c)** Iron and water temperature, both measured in pipe rack influent, exhibited comparable seasonal patterns to lead, the latter described by the seasonal term in the GAM.
@@
 
-The variation in dissolved lead evident in Figure 8a ignores seasonality by design.
+The variation in dissolved lead evident in Figure 8a neglects seasonality by design.
 And while the seasonal component of the model (Figure 8b) is described as a function of time (ordinal day), this is a proxy for the physical and chemical inputs that are expected to have caused it.
@@
 The decrease and subsequent increase in particulate lead beginning in late 2020 are not easily explained, but they may be related to water temperature.
-Seasonal temperature variation was most pronounced in the final 1.5 years of the study (Figure 8c), coinciding approximately with the apparent seasonal pattern in particulate lead (Figure 8a).
+Seasonal temperature variation was most pronounced in the final 1.5 years of the study (Figure 8c), coinciding approximately with an emerging apparent seasonal pattern in particulate lead over the same period (Figure 8a).
 Increasing feedwater aluminum concentrations may also have been a factor (Figure S9)---previous work has shown that aluminum and orthophosphate interact to increase particulate lead release [@trueman_seasonal_2022].
@@
 On the other hand, the initial 2 mg P L^-1^ dose caused excess lead release, highlighting an important trade-off that should be considered in applying a passivation-maintenance (high-low) dosing strategy.
-And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose may be a better option in this case.
-The effect of orthophosphate treatment was also complicated by pronounced seasonal variation in lead release, with a maximum in late summer and minimum in early spring.
-This seasonal pattern may have multiple drivers: iron and water temperature, for instance, were also seasonal.
+And since high orthophosphate doses may generate small lead-orthophosphate colloids/nanoparticles that are difficult to remove from drinking water [@lytle_lead_2020], the 1 mg P L^-1^ dose may be the better choice.
+The additional cost and downstream effects of elevated orthophosphate doses should also be considered: even in light of a thorough pipe loop study, water utilities face difficult decisions in applying orthophosphate.
 

### 2022-08-04 R/figures.R

@@ -395,8 +395,8 @@ fig3 <- gam_basis %>%
   filter(basis_name != "(Intercept)") %>%
   mutate(
     name = fct_recode(name,
-      "Unweighted basis functions of x" = "lpmat",
-      "Weighted basis functions of x" = "lpmat_wt",
+      "Unweighted basis functions of *t*, *s<sub>i</sub>(t)*" = "lpmat",
+      "Weighted basis functions of *t*, *s<sub>i</sub>(t)&beta;<sub>i</sub>*" = "lpmat_wt",
     )
   ) %>%
   ggplot(aes(date)) +
@@ -410,25 +410,26 @@ fig3 <- gam_basis %>%
         ungroup() %>%
         filter(!str_detect(name, "Unweighted"))
     },
-    aes(y = pred, col = "sum(italic(b[j]*'('*x[t]*')'*beta[j]), j == 1, k)"),
+    aes(y = pred, col = "sum(italic(s[i]*'('*t*')'*beta[i]), i == 1, n)"),
     size = 1
   ) +
   scale_color_manual(
     values = c(wesanderson::wes_palette("Zissou1", 9, type = "continuous"), "black"),
     labels = function(breaks) {
       breaks %>%
-        str_replace("s\\(numeric_date\\)\\.", "b") %>%
-        str_replace("(b)(\\d)", "italic(\\1[\\2]*'('*x[t]*')')") %>%
+        str_replace("s\\(numeric_date\\)\\.", "s") %>%
+        str_replace("(s)(\\d)", "italic(\\1[\\2]*'('*t*')')") %>%
         parse(text = .)
     }
   ) +
   theme(
     plot.margin = margin(r = .4, l = .1, b = .1, t = .1, unit = "cm"),
-    legend.text.align = 0
+    legend.text.align = 0,
+    strip.text = element_markdown()
   ) +
   labs(
     x = NULL,
-    y = expression("log([Pb] (µg L"^-1 * ")"),
+    y = "Basis function",
     col = NULL
   ) +
   guides(col = guide_legend(nrow = 4))

### 2022-08-04 R/render_paper.R

@@ -6,9 +6,9 @@ library("tidyverse")
 path <- "Rmarkdown/paper.Rmd"
 path_temp <- str_replace(path,"(.+/)(.+)", "\\1_\\2")
 
-read_file(path) %>% 
-  # move arabic numberal refs after punctuation:
-  str_replace_all("\\s+(\\[@.*?\\])([.,;:])?", "\\2\\1") %>% 
+read_file(path) %>%
+  # move arabic numeral refs after punctuation:
+  str_replace_all("\\s+(\\[@.*?\\])([.,;:])?", "\\2\\1") %>%
   write_file(path_temp)
 
 rmarkdown::render(path_temp)

### 2022-08-05 README.Rmd
update readme

@@
 
-This repository contains the R code and other materials for the paper "Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72-v3).
+This repository contains the R code and other materials for the paper "Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72-v4).
 Bespoke functions used to fit the models are available as a separate package, [`bgamcar1`](https://github.com/bentrueman/bgamcar1).

### 2022-09-16 Rmarkdown/paper.Rmd
pre-submission edits (revision 2)

@@
 
-Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting.
-They can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995].
+Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting model parameters.
+Nonlinearity can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995].
 The most common models for pipe loop data are usually inadequate to describe complex nonlinear variation [@simpson_modelling_2018].
@@
 The primary goal of a pipe loop study is to compare two or more corrosion control treatments [@wysock_statistical_1995].
-But the data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given their fairly rigid assumptions.
+But the data attributes outlined above can make it difficult to choose a statistical method from the suite that is commonly applied, given the necessary assumptions.
 Often, a more flexible approach is required, and Bayesian methods are inherently flexible since there is no need to derive the sampling distribution of an estimator [@kruschke_doing_2015].
@@
 Losses of dissolved lead by adsorption to PTFE filters have been documented previously [@minning_systematic_2015], which here would tend to bias dissolved concentrations low and particulate concentrations high.
-In hard waters such as the feedwater for the model system, this bias is likely to be much lower than in soft waters [@minning_systematic_2015].
+In hard waters such as the feedwater for the model system, though, this bias is likely to be much lower than in soft waters [@minning_systematic_2015].
 
@@
 And while the model assumes that seasonality strength does not vary over time, some previous work has suggested that orthophosphate may damp seasonal variation in lead [@colling_measurement_1987].
-This effect could be incorporated by allowing the seasonal smoother to vary with orthophosphate dose.
+This effect could be incorporated by allowing the seasonal smoother to vary with the orthophosphate dose.
 
@@
 \begin{align}
-likelihood: \\
+\text{likelihood}: \\
 y_t | censored_t = 0 \sim T(\mu_t,\sigma,\nu)~~[observed] \\ 
@@
 \\
-model~for~\mu_t: \\
+\text{model for }\mu_t\text{:} \\
 \mu_t = \alpha_{pipe_i} + \sum_{j=1}^{n}f_j(t) + \phi^s r_{t-s} \\
-priors: \\
+r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s) \\
+\\
+\text{priors}: \\
 \sigma \sim Half \text{-} T(0, 2.5, 3) \\
@@
 where $T$ represents the Student *t*-distribution with mean $\mu_t$, standard deviation $\sigma$, and degrees-of-freedom parameter $\nu$, determining the difference between the *t*-distribution and a Gaussian with the same $\mu$ and $\sigma$.
-($Gamma$ and $N$ represent the gamma and Gaussian distributions.)
+($Gamma$ and $N$ represent the gamma and Gaussian distributions, parameterized by the mean and shape parameter/standard deviation, respectively.)
 
@@
 $$
-The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$; $r_{t-s} = y_{t-s} - \alpha_{pipe_i} - \sum_{j=1}^{n}f_j(t-s)$.
+The autocorrelation coefficient $\phi^s$ is defined in equation $(4)$, and $r_{t-s}$ is the GAM residual at the previous time step, $t-s$.
 The $f_j(t)$ are smooth spline functions of time that comprise the GAM, and each of the $n$ smoothers takes the following form:
@@
 \begin{align}
-generalized~additive~model~(GAM): \\
+\text{generalized additive model (GAM)}: \\
 f_j(t) = X_j\beta_j + Z_jb_j \\
 \\
-priors~on~GAM~coefficients: \\
+\text{priors on GAM parameters}: \\
 \beta_j \sim T(0, 2.5, 3)~~[unpenalized] \\
 b_j \sim N(0, \sigma_b)~~[penalized] \\
-\sigma_b \sim Half \text{-} N(0, 0.5) (seasonal~smoother) \\
-\sigma_b \sim  Half \text{-} T(0, 2.5, 3) (global/group \text{-}level~smoothers) \\ 
+\sigma_b \sim Half \text{-} N(0, 0.5) [seasonal~smoother] \\
+\sigma_b \sim  Half \text{-} T(0, 2.5, 3) [global/group \text{-}level~smoothers] \\ 
 \end{align}
@@
 The stability of a pipe array---accounting for effluent from all pipes in the study---can be evaluated by considering the rate of change in this global term.
-Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized following commissioning (Figure 8a).
+Here, dissolved and particulate lead concentrations in pipe loop effluent declined significantly in the first year of operation as the system stabilized (Figure 8a).
 The subtler subsequent positive and negative trends in both fractions are difficult to explain, but they may be a response to changes in water quality not captured directly by the model (see *Trends in particulate lead release*).
@@
 The model is equally useful for other problems featuring the partially-censored, irregularly-sampled, autocorrelated time series common in water quality analysis [@helsel_statistical].
-But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real-time.
+But like many Bayesian models, this one is computationally-intensive to fit and would be difficult to implement in real time.
 

### 2022-09-20 Rmarkdown/paper.Rmd
resubmission to es&t eng

@@
 
-Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for water quality management.
+Pipe loop studies are used to evaluate corrosion control treatment, and updated regulatory guidance will ensure that they remain important for drinking water quality management.
 But the data they generate are difficult to analyze: non-linear time-trends, non-detects, extreme values, and autocorrelation are common attributes that make popular methods, such as the t- or rank-sum tests, poor descriptive models.
-Here, we propose a model for pipe loop data that accommodates many of these difficult-to-model attributes: a robust Bayesian generalized additive model with continuous-time autoregressive errors.
+Here, we propose a framework for describing pipe loop data that accommodates all of these challenging attributes: a robust Bayesian generalized additive model with continuous-time autoregressive errors.
 Our approach facilitates corrosion control treatment comparisons without the need for imputing non-detects or special handling of outliers.
@@
 
-Non-linear time trends and seasonal variation can be difficult to describe without over- or under-fitting model parameters.
+Non-linear time trends and seasonal variation can be challenging to describe without over- or under-fitting model parameters.
 Nonlinearity can also make it difficult to identify a stable phase, as typical guidance for analyzing pipe loop study data suggests [@wysock_statistical_1995].
@@
 ```{r figure-2, fig.cap="**Figure 2.
-(a)** Constant, linear, and generalized additive models (GAMs) fitted to an example (log-transformed) time series from the dataset, along with the residuals from each model.
+(a)** Constant, linear, and generalized additive models (GAM) fitted to an example (log-transformed) time series from the dataset, along with the residuals from each model.
 **(b)** Pearson correlation (*r*) between the residuals $\\epsilon$ at time *t* and *t-1*.
@@
 The generalized additive model (GAM) does better than the constant or even the linear model in accounting for autocorrelation in the data (Figure 2).
-This is because it accounts for non-linear patterns that the simpler models ignore [@trueman_seasonal_2022; @murphy_generalized_2019].
+This is because it describes non-linear patterns that the simpler models ignore [@trueman_seasonal_2022; @murphy_generalized_2019].
 GAMs can also be thought of as linear models, but instead of regressing y on x, we regress y on *functions* of x (cubic functions, for example, as in Figure 3).

### 2022-10-25 README.Rmd
readme update

@@
 
-This repository contains the R code and other materials for the paper "Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model", currently available as a [preprint](http://doi.org/10.26434/chemrxiv-2022-b3f72-v4).
+This repository contains the R code and other materials for the [paper](https://doi.org/10.1021/acsestengg.2c00194) "Comparing corrosion control treatments for drinking water using a robust Bayesian generalized additive model", also available as a [preprint](https://doi.org/10.26434/chemrxiv-2022-b3f72-v5).
 Bespoke functions used to fit the models are available as a separate package, [`bgamcar1`](https://github.com/bentrueman/bgamcar1).
